:::{#def-cond-pmf}

#### Conditional probability mass function

Let $X$ and $Y$ be [jointly distributed](random-variables.qmd#def-jointly-distributed) discrete random variables.
The **conditional probability mass function** of $Y$ given $X = x$
(for values of $x$ with $\P(X = x) > 0$) is:

$$\P(Y = y \mid X = x) \eqdef \frac{\P(X = x,\, Y = y)}{\P(X = x)}$$

:::

{{< slidebreak >}}

:::{#exm-cond-pmf}

#### Conditional PMF from a vaccine trial

The `vaccine` dataset in the `dobson` package
(@dobson4e, Table 9.6)
records the responses in a flu vaccine trial:
$X$ is treatment group (placebo or vaccine)
and $Y$ is the level of response to treatment (small, moderate, or large).

```{r}
#| code-fold: true
response_levels <- c("small", "moderate", "large")
vaccine_tab <-
  dobson::vaccine |>
  dplyr::mutate(response = factor(response, levels = response_levels)) |>
  xtabs(frequency ~ treatment + response, data = _)
pander::pander(as.data.frame.matrix(vaccine_tab))
n_vaccine <- sum(vaccine_tab)
n_placebo <- vaccine_tab["placebo", ]
```

The [joint PMF](random-variables.qmd#def-joint-pmf) of $(X, Y)$ is the table of frequencies divided by the total
$n = `r n_vaccine`$ trial participants.
By the [marginal PMF theorem](random-variables.qmd#thm-marginal-pmf),
the [marginal](random-variables.qmd#def-marginal) probability $\P(X = \text{placebo})$ is:

$$
\ba
\P(X = \text{placebo})
&= \P(X = \text{placebo},\, Y = \text{small})
 + \P(X = \text{placebo},\, Y = \text{moderate})
 + \P(X = \text{placebo},\, Y = \text{large})
&& \text{(marginal PMF from the joint PMF)}
\\&= \tfrac{`r n_placebo[["small"]]`}{`r n_vaccine`}
 + \tfrac{`r n_placebo[["moderate"]]`}{`r n_vaccine`}
 + \tfrac{`r n_placebo[["large"]]`}{`r n_vaccine`}
&& \text{(read the table)}
\\&= \tfrac{`r sum(n_placebo)`}{`r n_vaccine`}
&& \text{(add)}
\ea
$$

By @def-cond-pmf, the conditional PMF of $Y$ given $X = \text{placebo}$ is:

$$
\ba
\P(Y = \text{small} \mid X = \text{placebo})
&= \frac{\P(X = \text{placebo},\, Y = \text{small})}{\P(X = \text{placebo})}
&& \text{(definition of conditional PMF)}
\\&= \frac{`r n_placebo[["small"]]`/`r n_vaccine`}{`r sum(n_placebo)`/`r n_vaccine`}
&& \text{(substitute)}
\\&= \frac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
&& \text{(cancel } `r n_vaccine` \text{)}
\ea
$$

:::

{{< slidebreak >}}

:::{#def-cond-pdf}

#### Conditional probability density function

Let $X$ and $Y$ be jointly distributed continuous random variables
with [joint density](random-variables.qmd#def-joint-pdf) $\p(X = x,\, Y = y)$
and [marginal density](random-variables.qmd#thm-marginal-density) $\p(X = x)$.
The **conditional probability density function** of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

:::

{{< slidebreak >}}

:::{#exm-cond-pdf}

#### Conditional PDF from a bivariate normal model of birthweight data

The `birthweight` dataset in the `dobson` package
(@dobson4e, Table 2.3)
records gestational age (weeks) and birthweight (grams)
for 12 boys and 12 girls.
Let $X$ be gestational age and $Y$ be birthweight,
pooling both sexes into $n = 24$ observations.

```{r}
#| code-fold: true
birthweight <- dobson::birthweight
ga <- c(
  birthweight[["boys gestational age"]],
  birthweight[["girls gestational age"]]
)
wt <- c(birthweight[["boys weight"]], birthweight[["girls weight"]])
mu_x <- mean(ga)
sigma_x <- sd(ga)
mu_y <- mean(wt)
sigma_y <- sd(wt)
rho <- cor(ga, wt)
beta <- cov(ga, wt) / var(ga)
intercept <- mu_y - beta * mu_x
x0 <- 40
marg_dens_x0 <- dnorm(x0, mean = mu_x, sd = sigma_x)
cond_mean <- intercept + beta * x0
cond_sd <- sigma_y * sqrt(1 - rho^2)
tibble::tibble(
  parameter = c("mu_x", "sigma_x", "mu_y", "sigma_y", "rho"),
  estimate = round(c(mu_x, sigma_x, mu_y, sigma_y, rho), 4)
) |>
  pander::pander()
```

Modeling $(X, Y)$ as bivariate normal
with parameters set equal to these sample moments,
the joint density is
(a standard result; e.g. @CaseBerg01):

$$
\p(X=x,\,Y=y) = \frac{1}{2\pi\sdpar_X\sdpar_Y\sqrt{1-\corpar^2}}
\e{-\frac{1}{2(1-\corpar^2)}
\sb{\frac{(x-\mean_X)^2}{\varpar_X}
- \frac{2\corpar(x-\mean_X)(y-\mean_Y)}{\sdpar_X\sdpar_Y}
+ \frac{(y-\mean_Y)^2}{\varpar_Y}}}
$$

A further standard fact about the bivariate normal (@CaseBerg01)
is that the marginal distribution of $X$ is
[normal](random-variables.qmd#def-normal), $X \sim \ndist{\mean_X, \varpar_X}$.
At $x = `r x0`$ weeks (a full-term pregnancy),
$\mean_X = `r round(mu_x, 4)`$ and $\sdpar_X = `r round(sigma_x, 4)`$, so:

$$
\ba
\p(X=`r x0`) &= \frac{1}{\sdpar_X\sqrt{2\pi}}
  \e{-\frac{(`r x0`-\mean_X)^2}{2\varpar_X}}
&& \text{(normal density at } x = `r x0` \text{)}
\\&= \frac{1}{`r round(sigma_x, 4)`\sqrt{2\pi}}
  \e{-\frac{(`r x0`-`r round(mu_x, 4)`)^2}{2(`r round(sigma_x, 4)`)^2}}
&& \text{(substitute } \mean_X \text{ and } \sdpar_X \text{)}
\\&\approx `r round(marg_dens_x0, 4)`
&& \text{(evaluate)}
\ea
$$

By @def-cond-pdf, dividing the joint density by this marginal density
and simplifying the exponent
(completing the square in $y$; @CaseBerg01)
gives the conditional PDF of $Y$ given $X = `r x0`$,
which is itself normal
with mean shifted along the regression line
and variance reduced by a factor of $1-\corpar^2$:

$$
\ba
\p(Y=y \mid X=`r x0`) &= \frac{\p(X=`r x0`,\,Y=y)}{\p(X=`r x0`)}
&& \text{(definition of the conditional PDF)}
\\&= \frac{1}{\sdpar_Y\sqrt{2\pi(1-\corpar^2)}}
  \e{-\frac{1}{2(1-\corpar^2)}\sb{\frac{(`r x0`-\mean_X)^2}{\varpar_X}
  - \frac{2\corpar(`r x0`-\mean_X)(y-\mean_Y)}{\sdpar_X\sdpar_Y}
  + \frac{(y-\mean_Y)^2}{\varpar_Y}} + \frac{(`r x0`-\mean_X)^2}{2\varpar_X}}
&& \text{(substitute both densities; combine prefactors and exponents)}
\\&= \frac{1}{\sdpar_Y\sqrt{2\pi(1-\corpar^2)}}
  \e{-\frac{\corpar^2(`r x0`-\mean_X)^2}{2\varpar_X(1-\corpar^2)}
  + \frac{\corpar(`r x0`-\mean_X)(y-\mean_Y)}{\sdpar_X\sdpar_Y(1-\corpar^2)}
  - \frac{(y-\mean_Y)^2}{2\varpar_Y(1-\corpar^2)}}
&& \text{(combine the } (`r x0`-\mean_X)^2 \text{ terms: } 1 - \tfrac{1}{1-\corpar^2} = \tfrac{-\corpar^2}{1-\corpar^2} \text{)}
\\&= \frac{1}{\sdpar_Y\sqrt{2\pi(1-\corpar^2)}}
  \e{-\frac{1}{2\varpar_Y(1-\corpar^2)}
  \sb{\corpar^2\frac{\varpar_Y}{\varpar_X}(`r x0`-\mean_X)^2
  - 2\corpar\frac{\sdpar_Y}{\sdpar_X}(`r x0`-\mean_X)(y-\mean_Y)
  + (y-\mean_Y)^2}}
&& \text{(factor } -\tfrac{1}{2\varpar_Y(1-\corpar^2)} \text{ out of the exponent)}
\\&= \frac{1}{\sdpar_Y\sqrt{2\pi(1-\corpar^2)}}
  \e{-\frac{\paren{y - \sb{\mean_Y + \corpar\frac{\sdpar_Y}{\sdpar_X}(`r x0`-\mean_X)}}^2}{2\sdpar_Y^2(1-\corpar^2)}}
&& \text{(the bracket is a perfect square in } y \text{)}
\ea
$$

Multiplying out each of the last two exponents reproduces the one before it.

So $Y \mid X = `r x0` \sim \ndist{`r round(cond_mean, 2)`,\, `r round(cond_sd, 2)`^2}$:
the conditional mean, `r round(cond_mean, 2)` g,
is exactly the fitted regression line's prediction at $x=`r x0`$
($`r round(intercept, 4)` + `r round(beta, 4)` \times `r x0` = `r round(cond_mean, 2)`$),
matching `R`'s `lm(wt ~ ga)` fit directly:

```{r}
#| code-fold: true
pander::pander(coef(lm(wt ~ ga)))
```

:::

{{< slidebreak >}}

:::{#def-cond-expectation}

#### Conditional expectation

**Discrete case.**
Let $X$ and $Y$ be jointly distributed discrete random variables.
The **conditional expectation** of $Y$ given $X = x$,
using @def-cond-pmf,
is:

$$\E{Y \mid X = x} \eqdef \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)$$

**Continuous case.**
Let $X$ and $Y$ be jointly distributed continuous random variables.
The **conditional expectation** of $Y$ given $X = x$,
using @def-cond-pdf,
is:

$$\E{Y \mid X = x} \eqdef \int_{y \in \rangef{Y}} y \cd \p(Y = y \mid X = x)\, dy$$

:::

{{< slidebreak >}}

:::{#exm-cond-expectation}

#### Conditional expectation from real trial and birthweight data

**Discrete case.**
Continuing @exm-cond-pmf,
score the vaccine trial's response levels
$\text{small} = 1$, $\text{moderate} = 2$, $\text{large} = 3$.
The conditional PMF of $Y$ given $X = \text{placebo}$ (from @exm-cond-pmf) is:

$$
\ba
\P(Y = \text{small} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
\\
\P(Y = \text{moderate} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["moderate"]]`}{`r sum(n_placebo)`}
\\
\P(Y = \text{large} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["large"]]`}{`r sum(n_placebo)`}
\ea
$$

so:

$$
\ba
\E{Y \mid X = \text{placebo}}
&= 1 \cd \tfrac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
 + 2 \cd \tfrac{`r n_placebo[["moderate"]]`}{`r sum(n_placebo)`}
 + 3 \cd \tfrac{`r n_placebo[["large"]]`}{`r sum(n_placebo)`}
&& \text{(definition of conditional expectation)}
\\&= \frac{`r sum(1:3 * n_placebo)`}{`r sum(n_placebo)`}
&& \text{(common denominator)}
\\&\approx `r round(sum(1:3 * n_placebo) / sum(n_placebo), 2)`
&& \text{(divide)}
\ea
$$

**Continuous case.**
Continuing @exm-cond-pdf,
$Y \mid X = `r x0` \sim \ndist{`r round(cond_mean, 2)`,\, `r round(cond_sd, 2)`^2}$.
The mean of a normal distribution is its location parameter
(@CaseBerg01),
so integrating $y$ against this conditional density gives:

$$
\ba
\E{Y \mid X = `r x0`}
&= \int_{-\infty}^{\infty} y \cd \p(Y=y \mid X=`r x0`)\,dy
&& \text{(definition of conditional expectation)}
\\&= `r round(cond_mean, 2)` \text{ g}
&& \text{(the mean of a normal distribution is its location parameter)}
\ea
$$

matching the fitted regression line's prediction at $x=`r x0`$ weeks exactly,
as expected since the conditional mean of a bivariate normal
*is* the linear regression of $Y$ on $X$.

:::

{{< slidebreak >}}

:::{#def-cond-mixed}

#### Conditional expectation: mixed case

Suppose exactly one of $X, Y$ is discrete and the other is continuous,
with [joint density-mass function](random-variables.qmd#def-joint-density-mass) $\p(X = x,\, Y = y)$.

**$X$ discrete, $Y$ continuous.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $x$,
a probability density in $y$,
with $\int_{y \in \rangef{Y}} \p(X=x,\,Y=y)\,dy = \P(X=x)$.
The conditional PDF of $Y$ given $X = x$
(for values of $x$ with $\P(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\P(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \int_{y \in \rangef{Y}} y \cd \p(Y = y \mid X = x)\, dy$$

**$X$ continuous, $Y$ discrete.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $y$,
a probability density in $x$,
and $\sum_{y \in \rangef{Y}} \p(X=x,\,Y=y) = \p(X=x)$ is a density of $X$.
The conditional PMF of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\P(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)$$

:::

{{< slidebreak >}}

:::{#exm-cond-mixed}

#### Conditional expectation, one discrete variable and one continuous variable

**$X$ discrete, $Y$ continuous.**
The `plasma` dataset in the `dobson` package
(@dobson4e, Table 6.25)
records plasma inorganic phosphate levels (mg/dL)
one hour after a glucose tolerance test,
for hyperinsulinemic obese (`H-O`) and control (`C`) participants.
Let $X$ be group and $Y$ be phosphate level.

```{r}
#| code-fold: true
plasma_summary <-
  dobson::plasma |>
  dplyr::filter(Group %in% c("H-O", "C")) |>
  dplyr::summarise(
    .by = Group,
    n = dplyr::n(),
    mean = mean(phosphate),
    sd = sd(phosphate)
  )
group_stat <- function(group, stat) {
  plasma_summary[[stat]][plasma_summary[["Group"]] == group]
}
n_ho <- group_stat("H-O", "n")
n_c <- group_stat("C", "n")
mean_ho <- group_stat("H-O", "mean")
sd_ho <- group_stat("H-O", "sd")
mean_c <- group_stat("C", "mean")
sd_c <- group_stat("C", "sd")
pander::pander(plasma_summary)
```

Modeling phosphate as approximately normal within each group,
with parameters set equal to each group's sample mean and SD,
and $\P(X = \text{H-O}) = `r n_ho`/`r n_ho + n_c`$:

$$Y \mid X = \text{H-O} \sim \ndist{`r round(mean_ho, 2)`,\, `r round(sd_ho, 2)`^2}$$
$$Y \mid X = \text{C} \sim \ndist{`r round(mean_c, 2)`,\, `r round(sd_c, 2)`^2}$$

By @def-cond-mixed, since the mean of a normal distribution
is its location parameter:

$$\E{Y \mid X = \text{H-O}} = `r round(mean_ho, 2)` \text{ mg/dL}, \qquad
  \E{Y \mid X = \text{C}} = `r round(mean_c, 2)` \text{ mg/dL}$$

**$X$ continuous, $Y$ discrete.**
The `senility` dataset in the `dobson` package
(@dobson4e, Table 7.8)
records, for 54 elderly people,
a WAIS (Wechsler Adult Intelligence Scale) score
and whether symptoms of senility were present.
Let $X$ be WAIS score and $Y$ indicate senility symptoms.

```{r}
#| code-fold: true
senility_fit <- glm(s ~ x, data = dobson::senility, family = binomial)
b0 <- coef(senility_fit)[["(Intercept)"]]
b1 <- coef(senility_fit)[["x"]]
pander::pander(coef(senility_fit))
```

Modeling $\P(Y = 1 \mid X = x)$ with the fitted logistic regression:

$$\logit \P(Y = 1 \mid X = x) = `r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x$$

By @def-cond-mixed, since $Y \mid X=x$ is Bernoulli:

$$
\ba
\E{Y \mid X = x} &= 0 \cd \P(Y=0 \mid X=x) + 1 \cd \P(Y=1 \mid X=x)
&& \text{(definition of conditional expectation, mixed case)}
\\&= \P(Y=1 \mid X=x)
&& \text{(simplify)}
\\&= \frac{1}{1 + \e{-(`r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x)}}
&& \text{(invert the logit)}
\ea
$$

```{r}
#| code-fold: true
senility_p10 <- predict(
  senility_fit,
  newdata = tibble::tibble(x = 10),
  type = "response"
)
senility_p15 <- predict(
  senility_fit,
  newdata = tibble::tibble(x = 15),
  type = "response"
)
```

At $x = 10$: $\E{Y \mid X=10} = `r round(senility_p10, 2)`$.
At $x = 15$: $\E{Y \mid X=15} = `r round(senility_p15, 2)`$ ---
a lower WAIS score is associated with higher predicted probability
of senility symptoms.

:::

{{< slidebreak >}}

:::{#def-cond-expectation-function}

#### Conditional expectation function

The **conditional expectation function** $\E{Y \mid X}$ is the function
(and hence random variable) of $X$ obtained by evaluating
$\E{Y \mid X = x}$ at $X$; specifically,
$\E{Y \mid X} = g(X)$ where $g(x) \eqdef \E{Y \mid X = x}$.

:::

{{< slidebreak >}}

:::{#exm-cond-expectation-function}

#### The conditional expectation function of the birthweight model

Continuing @exm-cond-pdf and @exm-cond-expectation,
$(X, Y)$ (gestational age, birthweight) is modeled as bivariate normal.
The general form of the conditional mean derived in @exm-cond-pdf,
evaluated at an arbitrary $x$ instead of just $x=`r x0`$,
gives the conditional expectation function directly:

$$
\ba
g(x) &= \mean_Y + \corpar\frac{\sdpar_Y}{\sdpar_X}(x-\mean_X)
\\&= `r round(intercept, 4)` + `r round(beta, 4)`\, x
\ea
$$

which is exactly the fitted regression line
(the general algebraic identity
$\mean_Y + \corpar\frac{\sdpar_Y}{\sdpar_X}(x-\mean_X) = \paren{\mean_Y - \corpar\frac{\sdpar_Y}{\sdpar_X}\mean_X} + \corpar\frac{\sdpar_Y}{\sdpar_X}x$
is what makes the bivariate-normal conditional mean linear in $x$).
As a check, evaluating at $x = `r x0`$ recovers @exm-cond-expectation's result:

$$g(`r x0`) = `r round(intercept, 4)` + `r round(beta, 4)` \times `r x0` = `r round(intercept + beta*x0, 2)` \text{ g}$$

:::

{{< slidebreak >}}

:::{#def-cond-expectation-general}

#### Conditional expectation of a function of $X$ and $Y$

Let $X$ and $Y$ be jointly distributed random variables,
$h$ a function of two arguments,
and $x$ a value at which the conditional PMF or PDF of $Y$ given $X = x$ is defined
(@def-cond-pmf, @def-cond-pdf, or @def-cond-mixed).
The **conditional expectation** of $h(X, Y)$ given $X = x$ is,
for discrete $Y$:

$$\E{h(X, Y) \mid X = x} \eqdef \sum_{y \in \rangef{Y}} h(x, y) \cd \P(Y = y \mid X = x)$$

and for continuous $Y$:

$$\E{h(X, Y) \mid X = x} \eqdef \int_{y \in \rangef{Y}} h(x, y) \cd \p(Y = y \mid X = x)\,dy$$

when the sum or integral converges absolutely.
Evaluating this function of $x$ at $X$ gives the random variable $\E{h(X, Y) \mid X}$.

:::

{{< slidebreak >}}

:::{#exm-cond-expectation-general}

#### The second flip, given the first

In the [joint PMF of two coin flips](random-variables.qmd#exm-joint-pmf),
write $X$ for the first flip and $Y$ for the total number of heads.
Given $X = 1$, $Y$ is $1$ or $2$ with conditional probability $1/2$ each,
and $Y - X$ is the second flip.
With $h(x, y) = (y - x)^2$:

$$
\ba
\E{(Y - X)^2 \mid X = 1}
&= (1 - 1)^2 \cd \tfrac{1}{2} + (2 - 1)^2 \cd \tfrac{1}{2}
&& \text{(definition, with } x = 1 \text{)} \\
&= 0 + \tfrac{1}{2}
&& \text{(evaluate each term)} \\
&= \tfrac{1}{2}
&& \text{(add)}
\ea
$$

:::

{{< slidebreak >}}

:::{#cor-cond-expectation-general-y}
#### With $h(x, y) = y$, the general definition gives ${\E{Y \mid X = x}}$

For jointly distributed $X$ and $Y$ and $h(x, y) = y$,
@def-cond-expectation-general gives the conditional expectation of $Y$ given $X = x$
from @def-cond-expectation
(when $X$ and $Y$ are both discrete or both continuous)
or from @def-cond-mixed
(when one is discrete and the other continuous):

$$\E{h(X, Y) \mid X = x} = \E{Y \mid X = x}$$

:::

::: proof

For discrete $Y$:

$$
\ba
\E{h(X, Y) \mid X = x}
&= \sum_{y \in \rangef{Y}} h(x, y) \cd \P(Y = y \mid X = x)
&& \text{(conditional expectation of a function, discrete } Y \text{)} \\
&= \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)
&& \text{(} h(x, y) = y \text{)} \\
&= \E{Y \mid X = x}
&& \text{(conditional expectation, discrete } Y \text{)}
\ea
$$

The last step is @def-cond-expectation when $X$ is discrete
and @def-cond-mixed when $X$ is continuous.
For continuous $Y$, the same steps hold with integrals over $y$ in place of the sums
and $\p(Y = y \mid X = x)$ in place of $\P(Y = y \mid X = x)$.

:::

{{< slidebreak >}}

:::{#thm-cond-lotus}
#### Conditional LOTUS, discrete case

Let $X$ and $Y$ be jointly distributed discrete random variables,
$h$ a function of two arguments,
$W \eqdef h(X, Y)$,
and $x$ a value with $\P(X = x) > 0$.
Then $W$ is discrete,
and $\E{h(X, Y) \mid X = x}$ from @def-cond-expectation-general
equals $\E{W \mid X = x}$ from @def-cond-expectation,
with one sum converging absolutely exactly when the other does.

:::

::: proof

$W$ is discrete because its range is contained in
$\set{h(u, y) : u \in \rangef{X},\, y \in \rangef{Y}}$,
which is countable.
For each $w \in \rangef{W}$,
the event $\set{X = x,\, W = w}$ is the disjoint union
of the events $\set{X = x,\, Y = y}$ over the values $y \in \rangef{Y}$ with $h(x, y) = w$, so:

$$
\ba
\P(W = w \mid X = x)
&= \frac{\P(X = x,\, W = w)}{\P(X = x)}
&& \text{(conditional PMF of } W \text{)} \\
&= \frac{1}{\P(X = x)} \sum_{\substack{y \in \rangef{Y} \\ h(x, y) = w}} \P(X = x,\, Y = y)
&& \text{(countable additivity over the disjoint events } \set{X = x,\, Y = y} \text{)} \\
&= \sum_{\substack{y \in \rangef{Y} \\ h(x, y) = w}} \P(Y = y \mid X = x)
&& \text{(divide each term; conditional PMF of } Y \text{)}
\ea
$$

Every $y \in \rangef{Y}$ with $\P(Y = y \mid X = x) > 0$
has $h(x, y) \in \rangef{W}$,
because some outcome has $X = x$ and $Y = y$, and $W = h(x, y)$ there;
the other values of $y$ contribute terms equal to $0$.
So the groups $\set{y \in \rangef{Y} : h(x, y) = w}$, over $w \in \rangef{W}$,
contain every $y$ with a non-zero term, each exactly once:

$$
\ba
\E{W \mid X = x}
&= \sum_{w \in \rangef{W}} w \cd \P(W = w \mid X = x)
&& \text{(conditional expectation of } W \text{)} \\
&= \sum_{w \in \rangef{W}} w \cd \sum_{\substack{y \in \rangef{Y} \\ h(x, y) = w}} \P(Y = y \mid X = x)
&& \text{(the display above)} \\
&= \sum_{w \in \rangef{W}} \sum_{\substack{y \in \rangef{Y} \\ h(x, y) = w}} w \cd \P(Y = y \mid X = x)
&& \text{(distribute } w \text{ into the inner sum)} \\
&= \sum_{w \in \rangef{W}} \sum_{\substack{y \in \rangef{Y} \\ h(x, y) = w}} h(x, y) \cd \P(Y = y \mid X = x)
&& \text{(} w = h(x, y) \text{ for every term of the inner sum)} \\
&= \sum_{y \in \rangef{Y}} h(x, y) \cd \P(Y = y \mid X = x)
&& \text{(the groups contain every non-zero term once)} \\
&= \E{h(X, Y) \mid X = x}
&& \text{(conditional expectation of a function of } X \text{ and } Y \text{)}
\ea
$$

The same steps with $\abs{w}$ and $\abs{h(x, y)}$ in place of $w$ and $h(x, y)$
regroup a sum of non-negative terms, which is valid whether or not it converges,
so one sum converges absolutely exactly when the other does,
and then the regrouping above is valid too.

:::

::: {.remark .notes}

@def-cond-expectation-general is the conditional version of [LOTUS](#thm-lotus):
for discrete $X$, it is LOTUS applied under the probability measure $\Pr(\cdot \mid X = x)$,
and by @thm-cond-lotus, results about $\E{W \mid X}$ apply to it.
$Y$ can also be a pair $(Y, Z)$, with the sum or integral taken over both.

:::

{{< slidebreak >}}

:::{#thm-cond-linearity}

#### Linearity of conditional expectation

For jointly distributed $X$ and $Y$,
functions $h_1$ and $h_2$ whose conditional expectations given $X = x$ are defined,
and constants $a$ and $b$:

$$\E{a h_1(X, Y) + b h_2(X, Y) \mid X = x} = a\E{h_1(X, Y) \mid X = x} + b\E{h_2(X, Y) \mid X = x}$$

Evaluating both sides at $X$ gives
$\E{a h_1(X, Y) + b h_2(X, Y) \mid X} = a\E{h_1(X, Y) \mid X} + b\E{h_2(X, Y) \mid X}$.

:::

::: proof

For discrete $Y$:

$$
\ba
&\E{a h_1(X, Y) + b h_2(X, Y) \mid X = x}
\\&= \sum_{y \in \rangef{Y}} \paren{a h_1(x, y) + b h_2(x, y)} \cd \P(Y = y \mid X = x)
&& \text{(definition of conditional expectation)} \\
&= a \sum_{y \in \rangef{Y}} h_1(x, y) \cd \P(Y = y \mid X = x) + b \sum_{y \in \rangef{Y}} h_2(x, y) \cd \P(Y = y \mid X = x)
&& \text{(split the sum; factor out the constants)} \\
&= a\E{h_1(X, Y) \mid X = x} + b\E{h_2(X, Y) \mid X = x}
&& \text{(definition of conditional expectation)}
\ea
$$

Splitting the sum is valid because both sums converge absolutely.
For continuous $Y$, the same steps hold with integrals over $y$ in place of the sums,
by linearity of the integral:
unlike @thm-linearity-expectation, only one variable, $y$, is integrated,
against one conditional density.

:::

{{< slidebreak >}}

:::{#exm-cond-linearity}

#### Linearity, given the first flip

Continuing @exm-cond-expectation-general,
given $X = 1$, $Y$ is $1$ or $2$ with probability $1/2$ each,
so $\E{Y \mid X = 1} = 3/2$ and

$$
\begin{aligned}
\E{Y^2 \mid X = 1} &= (1 + 4)/2 \\
&= 5/2.
\end{aligned}
$$

By @thm-cond-linearity with

$$
\begin{aligned}
a &= b \\
&= 1:
\end{aligned}
$$

$$
\ba
\E{Y + Y^2 \mid X = 1}
&= \E{Y \mid X = 1} + \E{Y^2 \mid X = 1}
&& \text{(linearity of conditional expectation)} \\
&= \tfrac{3}{2} + \tfrac{5}{2}
&& \text{(substitute)} \\
&= 4
&& \text{(add)}
\ea
$$

As a check, directly:

$$
\begin{aligned}
(1 + 1) \cd \tfrac{1}{2} + (2 + 4) \cd \tfrac{1}{2} &= 1 + 3 \\
&= 4.
\end{aligned}
$$

:::

{{< slidebreak >}}

:::{#thm-cond-pull-out}

#### A function of $X$ factors out of a conditional expectation given $X$

For jointly distributed $X$ and $Y$,
a function $g$ of one argument,
and a function $h$ of two arguments
whose conditional expectation given $X = x$ is defined:

$$\E{g(X)\,h(X, Y) \mid X = x} = g(x) \cd \E{h(X, Y) \mid X = x}$$

Evaluating both sides at $X$ gives
$\E{g(X)\,h(X, Y) \mid X} = g(X) \cd \E{h(X, Y) \mid X}$.
In particular, taking $h = 1$ gives $\E{g(X) \mid X} = g(X)$.

:::

::: proof

For discrete $Y$:

$$
\ba
\E{g(X)\,h(X, Y) \mid X = x}
&= \sum_{y \in \rangef{Y}} g(x)\,h(x, y) \cd \P(Y = y \mid X = x)
&& \text{(definition of conditional expectation)} \\
&= g(x) \sum_{y \in \rangef{Y}} h(x, y) \cd \P(Y = y \mid X = x)
&& \text{(} g(x) \text{ does not depend on } y \text{)} \\
&= g(x) \cd \E{h(X, Y) \mid X = x}
&& \text{(definition of conditional expectation)}
\ea
$$

For $h = 1$, the remaining sum is $\sum_{y \in \rangef{Y}} \P(Y = y \mid X = x) = 1$:

$$
\ba
\sum_{y \in \rangef{Y}} \P(Y = y \mid X = x)
&= \sum_{y \in \rangef{Y}} \frac{\P(X = x,\, Y = y)}{\P(X = x)}
&& \text{(definition of the conditional PMF)} \\
&= \frac{\P(X = x)}{\P(X = x)}
&& \text{(marginal PMF from a joint PMF)} \\
&= 1
&& \text{(divide)}
\ea
$$

For continuous $Y$, the same steps hold with integrals over $y$ in place of the sums;
the conditional density integrates to 1 by
[the marginal density theorem](random-variables.qmd#thm-marginal-density)
(or, in the mixed case, by [the joint density-mass function's](random-variables.qmd#def-joint-density-mass) marginal identity).

:::

::: {.remark .notes}

This result is often summarized as "taking out what is known"
(e.g., @ross-probsim, Section 5.6.4):
given $X = x$, any function of $X$ is the known constant $g(x)$.

:::

{{< slidebreak >}}

:::{#exm-cond-pull-out}

#### Taking out the first flip

Continuing @exm-cond-linearity, with $g(x) = x$ and $h(x, y) = y$:

$$
\ba
\E{XY \mid X = 1}
&= 1 \cd \E{Y \mid X = 1}
&& \text{(a function of } X \text{ factors out)} \\
&= \tfrac{3}{2}
&& \text{(substitute } \E{Y \mid X = 1} = \tfrac{3}{2} \text{)}
\ea
$$

and

$$
\begin{aligned}
\E{XY \mid X = 0} &= 0 \cd \E{Y \mid X = 0} \\
&= 0.
\end{aligned}
$$

As a check, given $X = 1$ the product $XY$ equals $Y$, which is $1$ or $2$ with probability $1/2$ each.

:::
