:::{#def-cond-pmf}

#### Conditional probability mass function

Let $X$ and $Y$ be jointly distributed discrete random variables.
The **conditional probability mass function** of $Y$ given $X = x$
(for values of $x$ with $\P(X = x) > 0$) is:

$$\P(Y = y \mid X = x) \eqdef \frac{\P(X = x,\, Y = y)}{\P(X = x)}$$

:::

{{< slidebreak >}}

:::{#exm-cond-pmf}

#### Conditional PMF from a vaccine trial

The `vaccine` dataset in the `dobson` package
(@dobson4e, Table 9.6)
records the responses in a flu vaccine trial:
$X$ is treatment group (placebo or vaccine)
and $Y$ is the level of response to treatment (small, moderate, or large).

```{r}
#| code-fold: true
response_levels <- c("small", "moderate", "large")
vaccine_tab <-
  dobson::vaccine |>
  dplyr::mutate(response = factor(response, levels = response_levels)) |>
  xtabs(frequency ~ treatment + response, data = _)
pander::pander(as.data.frame.matrix(vaccine_tab))
n_vaccine <- sum(vaccine_tab)
n_placebo <- vaccine_tab["placebo", ]
```

The joint PMF of $(X, Y)$ is the table of frequencies divided by the total
$n = `r n_vaccine`$ trial participants.
The marginal probability $\P(X = \text{placebo})$ is:

$$
\ba
\P(X = \text{placebo})
&= \P(X = \text{placebo},\, Y = \text{small})
 + \P(X = \text{placebo},\, Y = \text{moderate})
 + \P(X = \text{placebo},\, Y = \text{large})
&& \text{(countable additivity)}
\\&= \tfrac{`r n_placebo[["small"]]`}{`r n_vaccine`}
 + \tfrac{`r n_placebo[["moderate"]]`}{`r n_vaccine`}
 + \tfrac{`r n_placebo[["large"]]`}{`r n_vaccine`}
&& \text{(read the table)}
\\&= \tfrac{`r sum(n_placebo)`}{`r n_vaccine`}
&& \text{(add)}
\ea
$$

By @def-cond-pmf, the conditional PMF of $Y$ given $X = \text{placebo}$ is:

$$
\ba
\P(Y = \text{small} \mid X = \text{placebo})
&= \frac{\P(X = \text{placebo},\, Y = \text{small})}{\P(X = \text{placebo})}
&& \text{(definition of conditional PMF)}
\\&= \frac{`r n_placebo[["small"]]`/`r n_vaccine`}{`r sum(n_placebo)`/`r n_vaccine`}
&& \text{(substitute)}
\\&= \frac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
&& \text{(cancel } `r n_vaccine` \text{)}
\ea
$$

:::

{{< slidebreak >}}

:::{#def-cond-pdf}

#### Conditional probability density function

Let $X$ and $Y$ be jointly distributed continuous random variables
with joint density $\p(X = x,\, Y = y)$ and marginal density $\p(X = x)$.
The **conditional probability density function** of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

:::

{{< slidebreak >}}

:::{#exm-cond-pdf}

#### Conditional PDF from a bivariate normal model of birthweight data

The `birthweight` dataset in the `dobson` package
(@dobson4e, Table 2.3)
records gestational age (weeks) and birthweight (grams)
for 12 boys and 12 girls.
Let $X$ be gestational age and $Y$ be birthweight,
pooling both sexes into $n = 24$ observations.

```{r}
#| code-fold: true
birthweight <- dobson::birthweight
ga <- c(
  birthweight[["boys gestational age"]],
  birthweight[["girls gestational age"]]
)
wt <- c(birthweight[["boys weight"]], birthweight[["girls weight"]])
mu_x <- mean(ga)
sigma_x <- sd(ga)
mu_y <- mean(wt)
sigma_y <- sd(wt)
rho <- cor(ga, wt)
beta <- cov(ga, wt) / var(ga)
intercept <- mu_y - beta * mu_x
x0 <- 40
marg_dens_x0 <- dnorm(x0, mean = mu_x, sd = sigma_x)
cond_mean <- intercept + beta * x0
cond_sd <- sigma_y * sqrt(1 - rho^2)
pander::pander(data.frame(
  parameter = c("mu_x", "sigma_x", "mu_y", "sigma_y", "rho"),
  estimate = round(c(mu_x, sigma_x, mu_y, sigma_y, rho), 4)
))
```

Modeling $(X, Y)$ as bivariate normal
with parameters set equal to these sample moments,
the joint density is
(a standard result; e.g. @CaseBerg01):

$$
\p(X=x,\,Y=y) = \frac{1}{2\pi\sigma_X\sigma_Y\sqrt{1-\rho^2}}
\e{-\frac{1}{2(1-\rho^2)}
\sb{\frac{(x-\mu_X)^2}{\sigma_X^2}
- \frac{2\rho(x-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y}
+ \frac{(y-\mu_Y)^2}{\sigma_Y^2}}}
$$

A further standard fact about the bivariate normal (@CaseBerg01)
is that the marginal distribution of $X$ is
$X \sim \ndist{\mu_X, \sigma_X^2}$.
At $x = `r x0`$ weeks (a full-term pregnancy),
$\mu_X = `r round(mu_x, 4)`$ and $\sigma_X = `r round(sigma_x, 4)`$, so:

$$
\ba
\p(X=`r x0`) &= \frac{1}{\sigma_X\sqrt{2\pi}}
  \e{-\frac{(`r x0`-\mu_X)^2}{2\sigma_X^2}}
\\&= \frac{1}{`r round(sigma_x, 4)`\sqrt{2\pi}}
  \e{-\frac{(`r x0`-`r round(mu_x, 4)`)^2}{2(`r round(sigma_x, 4)`)^2}}
\\&\approx `r round(marg_dens_x0, 4)`
\ea
$$

By @def-cond-pdf, dividing the joint density by this marginal density
and simplifying the exponent
(completing the square in $y$; @CaseBerg01)
gives the conditional PDF of $Y$ given $X = `r x0`$,
which is itself normal
with mean shifted along the regression line
and variance reduced by a factor of $1-\rho^2$:

$$
\ba
\p(Y=y \mid X=`r x0`) &= \frac{\p(X=`r x0`,\,Y=y)}{\p(X=`r x0`)}
\\&= \frac{1}{\sigma_Y\sqrt{2\pi(1-\rho^2)}}
  \e{-\frac{1}{2(1-\rho^2)}\sb{\frac{(`r x0`-\mu_X)^2}{\sigma_X^2}
  - \frac{2\rho(`r x0`-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y}
  + \frac{(y-\mu_Y)^2}{\sigma_Y^2}} + \frac{(`r x0`-\mu_X)^2}{2\sigma_X^2}}
\\&= \frac{1}{\sigma_Y\sqrt{2\pi(1-\rho^2)}}
  \e{-\frac{\rho^2(`r x0`-\mu_X)^2}{2\sigma_X^2(1-\rho^2)}
  + \frac{\rho(`r x0`-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y(1-\rho^2)}
  - \frac{(y-\mu_Y)^2}{2\sigma_Y^2(1-\rho^2)}}
\\&= \frac{1}{\sigma_Y\sqrt{2\pi(1-\rho^2)}}
  \e{-\frac{1}{2\sigma_Y^2(1-\rho^2)}
  \sb{\rho^2\frac{\sigma_Y^2}{\sigma_X^2}(`r x0`-\mu_X)^2
  - 2\rho\frac{\sigma_Y}{\sigma_X}(`r x0`-\mu_X)(y-\mu_Y)
  + (y-\mu_Y)^2}}
\\&= \frac{1}{\sigma_Y\sqrt{2\pi(1-\rho^2)}}
  \e{-\frac{\paren{y - \sb{\mu_Y + \rho\frac{\sigma_Y}{\sigma_X}(`r x0`-\mu_X)}}^2}{2\sigma_Y^2(1-\rho^2)}}
\ea
$$

The first equality substitutes the joint and marginal densities
and combines their prefactors and exponents;
the second equality combines the two $(`r x0`-\mu_X)^2$ terms
using $1 - \tfrac{1}{1-\rho^2} = \tfrac{-\rho^2}{1-\rho^2}$;
the third equality factors $-\tfrac{1}{2\sigma_Y^2(1-\rho^2)}$
out of the bracket
(multiplying through confirms it reproduces the previous line);
and the fourth equality completes the square in $y$,
since expanding
$\paren{y - \sb{\mu_Y + \rho\frac{\sigma_Y}{\sigma_X}(`r x0`-\mu_X)}}^2$
reproduces exactly the bracketed quadratic in the third equality.

So $Y \mid X = `r x0` \sim \ndist{`r round(cond_mean, 2)`,\, `r round(cond_sd, 2)`^2}$:
the conditional mean, `r round(cond_mean, 2)` g,
is exactly the fitted regression line's prediction at $x=`r x0`$
($`r round(intercept, 4)` + `r round(beta, 4)` \times `r x0` = `r round(cond_mean, 2)`$),
matching `R`'s `lm(wt ~ ga)` fit directly:

```{r}
#| code-fold: true
pander::pander(coef(lm(wt ~ ga)))
```

:::

{{< slidebreak >}}

:::{#def-cond-expectation}

#### Conditional expectation

**Discrete case.**
Let $X$ and $Y$ be jointly distributed discrete random variables.
The **conditional expectation** of $Y$ given $X = x$,
using @def-cond-pmf,
is:

$$\E{Y \mid X = x} \eqdef \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)$$

**Continuous case.**
Let $X$ and $Y$ be jointly distributed continuous random variables.
The **conditional expectation** of $Y$ given $X = x$,
using @def-cond-pdf,
is:

$$\E{Y \mid X = x} \eqdef \int_{y \in \rangef{Y}} y \cd \p(Y = y \mid X = x)\, dy$$

:::

{{< slidebreak >}}

:::{#exm-cond-expectation}

#### Conditional expectation from real trial and birthweight data

**Discrete case.**
Continuing @exm-cond-pmf,
score the vaccine trial's response levels
$\text{small} = 1$, $\text{moderate} = 2$, $\text{large} = 3$.
The conditional PMF of $Y$ given $X = \text{placebo}$ (from @exm-cond-pmf) is:

$$
\ba
\P(Y = \text{small} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
\\
\P(Y = \text{moderate} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["moderate"]]`}{`r sum(n_placebo)`}
\\
\P(Y = \text{large} \mid X = \text{placebo}) &= \tfrac{`r n_placebo[["large"]]`}{`r sum(n_placebo)`}
\ea
$$

so:

$$
\ba
\E{Y \mid X = \text{placebo}}
&= 1 \cd \tfrac{`r n_placebo[["small"]]`}{`r sum(n_placebo)`}
 + 2 \cd \tfrac{`r n_placebo[["moderate"]]`}{`r sum(n_placebo)`}
 + 3 \cd \tfrac{`r n_placebo[["large"]]`}{`r sum(n_placebo)`}
&& \text{(definition of conditional expectation)}
\\&= \frac{`r sum(1:3 * n_placebo)`}{`r sum(n_placebo)`}
&& \text{(common denominator)}
\\&\approx `r round(sum(1:3 * n_placebo) / sum(n_placebo), 2)`
&& \text{(divide)}
\ea
$$

**Continuous case.**
Continuing @exm-cond-pdf,
$Y \mid X = `r x0` \sim \ndist{`r round(cond_mean, 2)`,\, `r round(cond_sd, 2)`^2}$.
The mean of a normal distribution is its location parameter
(@CaseBerg01),
so integrating $y$ against this conditional density gives:

$$
\ba
\E{Y \mid X = `r x0`}
&= \int_{-\infty}^{\infty} y \cd \p(Y=y \mid X=`r x0`)\,dy
\\&= `r round(cond_mean, 2)` \text{ g}
\ea
$$

matching the fitted regression line's prediction at $x=`r x0`$ weeks exactly,
as expected since the conditional mean of a bivariate normal
*is* the linear regression of $Y$ on $X$.

:::

{{< slidebreak >}}

:::{#def-cond-mixed}

#### Conditional expectation: mixed case

Suppose exactly one of $X, Y$ is discrete and the other is continuous.
Write $\p(X = x,\, Y = y)$ for their **joint density-mass function**:
a probability density in the continuous variable
and a probability mass in the discrete variable.

**$X$ discrete, $Y$ continuous.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $x$,
a probability density in $y$,
scaled so that $\int_{y} \p(X=x,\,Y=y)\,dy = \P(X=x)$.
The conditional PDF of $Y$ given $X = x$
(for values of $x$ with $\P(X = x) > 0$) is:

$$\p(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\P(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \int_{y \in \rangef{Y}} y \cd \p(Y = y \mid X = x)\, dy$$

**$X$ continuous, $Y$ discrete.**
Here $\p(X=x,\,Y=y)$ is, for each fixed $y$,
a probability density in $x$,
scaled so that $\sum_{y} \p(X=x,\,Y=y) = \p(X=x)$.
The conditional PMF of $Y$ given $X = x$
(for values of $x$ with $\p(X = x) > 0$) is:

$$\P(Y = y \mid X = x) \eqdef \frac{\p(X = x,\, Y = y)}{\p(X = x)}$$

and the conditional expectation of $Y$ given $X = x$ is:

$$\E{Y \mid X = x} \eqdef \sum_{y \in \rangef{Y}} y \cd \P(Y = y \mid X = x)$$

:::

{{< slidebreak >}}

:::{#exm-cond-mixed}

#### Conditional expectation, one discrete variable and one continuous variable

**$X$ discrete, $Y$ continuous.**
The `plasma` dataset in the `dobson` package
(@dobson4e, Table 6.25)
records plasma inorganic phosphate levels (mg/dL)
one hour after a glucose tolerance test,
for hyperinsulinemic obese (`H-O`) and control (`C`) participants.
Let $X$ be group and $Y$ be phosphate level.

```{r}
#| code-fold: true
plasma_summary <-
  dobson::plasma |>
  dplyr::filter(Group %in% c("H-O", "C")) |>
  dplyr::summarise(
    .by = Group,
    n = dplyr::n(),
    mean = mean(phosphate),
    sd = sd(phosphate)
  )
group_stat <- function(group, stat) {
  plasma_summary[[stat]][plasma_summary[["Group"]] == group]
}
n_ho <- group_stat("H-O", "n")
n_c <- group_stat("C", "n")
mean_ho <- group_stat("H-O", "mean")
sd_ho <- group_stat("H-O", "sd")
mean_c <- group_stat("C", "mean")
sd_c <- group_stat("C", "sd")
pander::pander(plasma_summary)
```

Modeling phosphate as approximately normal within each group,
with parameters set equal to each group's sample mean and SD,
and $\P(X = \text{H-O}) = `r n_ho`/`r n_ho + n_c`$:

$$Y \mid X = \text{H-O} \sim \ndist{`r round(mean_ho, 2)`,\, `r round(sd_ho, 2)`^2}$$
$$Y \mid X = \text{C} \sim \ndist{`r round(mean_c, 2)`,\, `r round(sd_c, 2)`^2}$$

By @def-cond-mixed, since the mean of a normal distribution
is its location parameter:

$$\E{Y \mid X = \text{H-O}} = `r round(mean_ho, 2)` \text{ mg/dL}, \qquad
  \E{Y \mid X = \text{C}} = `r round(mean_c, 2)` \text{ mg/dL}$$

**$X$ continuous, $Y$ discrete.**
The `senility` dataset in the `dobson` package
(@dobson4e, Table 7.8)
records, for 54 elderly people,
a WAIS (Wechsler Adult Intelligence Scale) score
and whether symptoms of senility were present.
Let $X$ be WAIS score and $Y$ indicate senility symptoms.

```{r}
#| code-fold: true
senility_fit <- glm(s ~ x, data = dobson::senility, family = binomial)
b0 <- coef(senility_fit)[["(Intercept)"]]
b1 <- coef(senility_fit)[["x"]]
pander::pander(coef(senility_fit))
```

Modeling $\P(Y = 1 \mid X = x)$ with the fitted logistic regression:

$$\logit \P(Y = 1 \mid X = x) = `r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x$$

By @def-cond-mixed, since $Y \mid X=x$ is Bernoulli:

$$
\ba
\E{Y \mid X = x} &= 0 \cd \P(Y=0 \mid X=x) + 1 \cd \P(Y=1 \mid X=x)
\\&= \P(Y=1 \mid X=x)
\\&= \frac{1}{1 + \e{-(`r sprintf("%.4f", b0)` - `r sprintf("%.4f", abs(b1))`\, x)}}
\ea
$$

```{r}
#| code-fold: true
senility_p10 <- predict(
  senility_fit,
  newdata = data.frame(x = 10),
  type = "response"
)
senility_p15 <- predict(
  senility_fit,
  newdata = data.frame(x = 15),
  type = "response"
)
```

At $x = 10$: $\E{Y \mid X=10} = `r round(senility_p10, 2)`$.
At $x = 15$: $\E{Y \mid X=15} = `r round(senility_p15, 2)`$ ---
a lower WAIS score is associated with higher predicted probability
of senility symptoms.

:::

{{< slidebreak >}}

:::{#def-cond-expectation-function}

#### Conditional expectation function

The **conditional expectation function** $\E{Y \mid X}$ is the function
(and hence random variable) of $X$ obtained by evaluating
$\E{Y \mid X = x}$ at $X$; specifically,
$\E{Y \mid X} = g(X)$ where $g(x) \eqdef \E{Y \mid X = x}$.

:::

{{< slidebreak >}}

:::{#exm-cond-expectation-function}

#### The conditional expectation function of the birthweight model

Continuing @exm-cond-pdf and @exm-cond-expectation,
$(X, Y)$ (gestational age, birthweight) is modeled as bivariate normal.
The general form of the conditional mean derived in @exm-cond-pdf,
evaluated at an arbitrary $x$ instead of just $x=`r x0`$,
gives the conditional expectation function directly:

$$
\ba
g(x) &= \mu_Y + \rho\frac{\sigma_Y}{\sigma_X}(x-\mu_X)
\\&= `r round(intercept, 4)` + `r round(beta, 4)`\, x
\ea
$$

which is exactly the fitted regression line
(the general algebraic identity
$\mu_Y + \rho\frac{\sigma_Y}{\sigma_X}(x-\mu_X) = \paren{\mu_Y - \rho\frac{\sigma_Y}{\sigma_X}\mu_X} + \rho\frac{\sigma_Y}{\sigma_X}x$
is what makes the bivariate-normal conditional mean linear in $x$).
As a check, evaluating at $x = `r x0`$ recovers @exm-cond-expectation's result:

$$g(`r x0`) = `r round(intercept, 4)` + `r round(beta, 4)` \times `r x0` = `r round(intercept + beta*x0, 2)` \text{ g}$$

:::
