:::{#exm-laplace-vs-normal}
#### Laplace and normal tails with the same mean and variance

A Laplace variable $Y \sim \dLaplacef{0, b}$ has variance $2b^2$ (@thm-laplace-moments).
Choosing $b = 1/\sqrt{2}$ gives variance $1$,
the same mean and variance as a standard normal.
@fig-laplace-vs-normal compares the two densities.
The Laplace density is higher at the center and in the far tails,
and lower in between.

```{r}
#| label: fig-laplace-vs-normal
#| fig-cap: "Laplace and standard normal densities with the same mean (0) and variance (1)"
#| fig-alt: >-
#|   Two symmetric curves over y from minus 4 to 4. The standard normal is a
#|   smooth bell. The Laplace curve with the same variance peaks higher at 0
#|   with a sharp point, falls below the normal curve at moderate distances,
#|   and stays above it in the far tails.
#| code-fold: true
b <- 1 / sqrt(2)
grid <- seq(-4, 4, length.out = 801)
dlaplace <- function(y, mu = 0, b = 1) exp(-abs(y - mu) / b) / (2 * b)
densities <- tibble::tibble(
  y = rep(grid, times = 2),
  distribution = rep(c("Laplace", "Normal"), each = length(grid)),
  density = c(dlaplace(grid, b = b), dnorm(grid))
)
ggplot2::ggplot(densities) +
  ggplot2::aes(x = y, y = density, colour = distribution) +
  ggplot2::geom_line() +
  ggplot2::labs(colour = NULL)
```

The probability of landing more than $3$ standard deviations from the mean shows the difference in the tails.
For the Laplace distribution,
$\P(\abs{Y} > t) = \ef{-t/b}$,
because each tail has probability $\frac{1}{2}\ef{-t/b}$ by integrating @eq-laplace-pdf.

```{r}
#| label: laplace-vs-normal-tails
#| code-fold: false
t <- 3
c(
  Laplace = exp(-t / b),
  Normal = 2 * pnorm(-t)
)
```

The Laplace distribution puts about `r round(exp(-3 / b) / (2 * pnorm(-3)), 1)` times as much probability beyond $3$ standard deviations as the normal distribution does.
:::
