{{< slidebreak >}}

:::{#exm-clt-auto}
#### Sample means of car horsepower

The Auto data record the horsepower and fuel economy of
`{r} nrow(read.csv("data/auto.csv"))` car models
sold between 1970 and 1982.
They come from the StatLib library at Carnegie Mellon University
and ship with the book *An Introduction to Statistical Learning*,
in the Python package [`ISLP`](https://islp.readthedocs.io/)
(`ISLP.load_data("Auto")`)
and the R package [`ISLR2`](https://cran.r-project.org/package=ISLR2)
(`ISLR2::Auto`).
This site has a copy at [`data/auto.csv`](data/auto.csv),
and both code tabs below read that file.

Horsepower is not bell-shaped:
a few powerful cars give it a long right tail.
To apply @thm-clt, treat each draw of one car at random as one IID observation $X_i$,
so that $\mean$ and $\varpar$ are the mean and variance of the 392 horsepower values.
The sum $S_n$ has mean $n\mean$ and variance $n\varpar$
by @cor-sum-iid-moments,
so the sample mean $S_n/n$ has mean $\mean$ and standard deviation $\sdpar/\sqrt{n}$.
The code below draws 10,000 random samples of each size $n \in \set{1, 5, 30}$,
with replacement, and computes each sample's mean.

::: {.panel-tabset}

#### Python

```{python}
#| code-fold: false
import numpy as np
import pandas as pd

hp = pd.read_csv("data/auto.csv")["horsepower"].to_numpy(dtype=float)
mu, sigma = hp.mean(), hp.std(ddof=0)

rng = np.random.default_rng(571)
sizes = [1, 5, 30]
sample_means = {n: rng.choice(hp, size=(10_000, n)).mean(axis=1) for n in sizes}

def skewness(x):
    return np.mean(((x - x.mean()) / x.std()) ** 3)

pd.DataFrame({
    "n": sizes,
    "sd of sample means": [sample_means[n].std(ddof=1) for n in sizes],
    "sigma / sqrt(n)": [sigma / np.sqrt(n) for n in sizes],
    "skewness of sample means": [skewness(sample_means[n]) for n in sizes],
})
```

#### R

```{r}
#| label: clt-auto-r
#| code-fold: false
hp <- read.csv("data/auto.csv")$horsepower
mu <- mean(hp)
sigma <- sqrt(mean((hp - mu)^2))

set.seed(571)
sizes <- c(1, 5, 30)
sample_means <- lapply(sizes, function(n) {
  replicate(10000, mean(sample(hp, n, replace = TRUE)))
})

skewness <- function(x) mean(((x - mean(x)) / sqrt(mean((x - mean(x))^2)))^3)

data.frame(
  n = sizes,
  sd_of_sample_means = sapply(sample_means, sd),
  sigma_over_root_n = sigma / sqrt(sizes),
  skewness_of_sample_means = sapply(sample_means, skewness)
)
```

:::

```{r}
#| include: false
# The claims the prose makes: both languages see the same population, the
# simulated spread matches sigma / sqrt(n), and the normal approximation
# improves with n.
stopifnot(
  abs(mu - reticulate::py$mu) < 1e-9,
  abs(sigma - reticulate::py$sigma) < 1e-9
)
sd_ratio <- sapply(seq_along(sizes), function(k) {
  sd(sample_means[[k]]) / (sigma / sqrt(sizes[k]))
})
skew_of_means <- sapply(sample_means, skewness)
stopifnot(
  all(abs(sd_ratio - 1) < 0.05),
  skew_of_means[1] > 0.5,
  abs(skew_of_means[3]) < abs(skew_of_means[2]),
  abs(skew_of_means[2]) < abs(skew_of_means[1])
)
```

@fig-clt-auto shows a histogram of the 10,000 sample means for each $n$,
with the normal density $\ndist{\mean, \varpar/n}$ from @thm-clt drawn over it.

```{r}
#| label: fig-clt-auto
#| fig-cap: >-
#|   Histograms of 10,000 sample means of horsepower, for samples of
#|   n = 1, 5 and 30 cars drawn with replacement from the Auto data.
#|   The curve is the normal density with mean mu and standard
#|   deviation sigma over the square root of n.
#| fig-alt: >-
#|   Three histograms side by side. The first, for one car, is skewed
#|   to the right with a long tail. The second, for five cars, is
#|   less skewed. The third, for thirty cars, is symmetric and
#|   bell-shaped. A normal curve drawn over each fits the third
#|   closely and the first poorly.
#| fig-height: 3
#| fig-width: 8
#| code-fold: true
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (k in seq_along(sizes)) {
  hist(sample_means[[k]], breaks = 40, freq = FALSE,
       main = paste("n =", sizes[k]), xlab = "sample mean of horsepower")
  curve(dnorm(x, mu, sigma / sqrt(sizes[k])), add = TRUE, lwd = 2)
}
par(op)
```

The sample means for $n = 1$ are just single cars, so they are skewed.
By $n = 30$ the histogram is close to the normal curve.
The standard deviation of the sample means is within
`{r} round(100 * max(abs(sd_ratio - 1)), 1)`%
of $\sdpar/\sqrt{n}$ at every $n$,
as @cor-sum-iid-moments predicts.
The skewness of the sample means,
which is zero for a normal distribution,
falls from `{r} round(skew_of_means[1], 2)` at $n = 1$
to `{r} round(skew_of_means[2], 2)` at $n = 5$
and `{r} round(skew_of_means[3], 2)` at $n = 30$.

:::
