{{< slidebreak >}}

:::{#exm-cor-auto}
#### Correlation of horsepower and fuel economy

The Auto data record the horsepower and fuel economy (miles per gallon)
of `{r} nrow(read.csv("data/auto.csv"))` car models
sold between 1970 and 1982.
They come from the StatLib library at Carnegie Mellon University
and ship with the book *An Introduction to Statistical Learning*,
in the Python package [`ISLP`](https://islp.readthedocs.io/)
(`ISLP.load_data("Auto")`)
and the R package [`ISLR2`](https://cran.r-project.org/package=ISLR2)
(`ISLR2::Auto`).
This site has a copy at [`data/auto.csv`](data/auto.csv),
and both code tabs below read that file.

Choose one car at random, each with the same probability,
and let $X$ be its horsepower and $Y$ its miles per gallon.
Then $\E{X}$, $\Var{X}$, and $\Cov{X,Y}$ are the averages over the cars
in the data, and @def-correlation gives $\Cor{X,Y}$.
The code below computes each part of the definition directly.

::: {.panel-tabset}

#### Python

```{python}
#| code-fold: false
import numpy as np
import pandas as pd

auto = pd.read_csv("data/auto.csv")
x = auto["horsepower"].to_numpy(dtype=float)
y = auto["mpg"].to_numpy(dtype=float)

cov_xy = np.mean((x - x.mean()) * (y - y.mean()))
sd_x = np.sqrt(np.mean((x - x.mean()) ** 2))
sd_y = np.sqrt(np.mean((y - y.mean()) ** 2))

pd.Series({
    "Cov(X, Y)": cov_xy,
    "SD(X)": sd_x,
    "SD(Y)": sd_y,
    "Cor(X, Y)": cov_xy / (sd_x * sd_y),
})
```

#### R

```{r}
#| label: cor-auto-r
#| code-fold: false
auto <- read.csv("data/auto.csv")
x <- auto$horsepower
y <- auto$mpg

cov_xy <- mean((x - mean(x)) * (y - mean(y)))
sd_x <- sqrt(mean((x - mean(x))^2))
sd_y <- sqrt(mean((y - mean(y))^2))

c(
  "Cov(X, Y)" = cov_xy,
  "SD(X)" = sd_x,
  "SD(Y)" = sd_y,
  "Cor(X, Y)" = cov_xy / (sd_x * sd_y)
)
```

:::

```{r}
#| include: false
# The claims the prose makes: both languages agree, the definition matches
# R's built-in sample correlation, and the correlation is negative.
stopifnot(
  abs(cov_xy - reticulate::py$cov_xy) < 1e-9,
  abs(cov_xy / (sd_x * sd_y) - cor(x, y)) < 1e-9,
  cov_xy < 0
)
```

@fig-cor-auto plots each car's miles per gallon against its horsepower.

```{r}
#| label: fig-cor-auto
#| fig-cap: >-
#|   Miles per gallon against horsepower for the cars in the Auto data.
#|   Cars with more horsepower tend to have lower fuel economy.
#| fig-alt: >-
#|   A scatter plot of miles per gallon on the vertical axis against
#|   horsepower on the horizontal axis. The points slope downward from
#|   the upper left to the lower right and curve toward a flat floor.
#| fig-height: 3.5
#| fig-width: 5
#| code-fold: true
plot(x, y, xlab = "horsepower", ylab = "miles per gallon", pch = 16,
     col = adjustcolor("black", alpha.f = 0.4))
```

The covariance is
`{r} round(cov_xy, 1)`,
in units of horsepower times miles per gallon.
It is negative, so cars with more horsepower than average tend to have
lower fuel economy than average.
Dividing by the two standard deviations,
`{r} round(sd_x, 1)` horsepower and `{r} round(sd_y, 1)` miles per gallon,
removes the units and gives a correlation of
`{r} round(cov_xy / (sd_x * sd_y), 2)`,
which lies between $-1$ and $1$ as @thm-cor-bounds requires.
The same value comes from R's built-in `cor(x, y)`,
because the sample correlation divides the covariance and both standard
deviations by the same count, which cancels.

:::
