{{< slidebreak >}}

:::{#exm-lie-auto}
#### Fuel economy by number of cylinders

The Auto data record the number of engine cylinders and the fuel economy
(miles per gallon) of
`{r} nrow(read.csv("data/auto.csv"))` car models
sold between 1970 and 1982.
They come from the StatLib library at Carnegie Mellon University
and ship with the book *An Introduction to Statistical Learning*,
in the Python package [`ISLP`](https://islp.readthedocs.io/)
(`ISLP.load_data("Auto")`)
and the R package [`ISLR2`](https://cran.r-project.org/package=ISLR2)
(`ISLR2::Auto`).
This site has a copy at [`data/auto.csv`](data/auto.csv),
and both code tabs below read that file.

Choose one car at random, each with the same probability,
and let $X$ be its number of cylinders and $Y$ its miles per gallon.
Then $\P(X = x)$ is the fraction of cars with $x$ cylinders,
and $\E{Y \mid X = x}$ from @def-cond-expectation is the average
miles per gallon among those cars.
The code below computes both for every $x$,
then checks @thm-lie by averaging the conditional means
with weights $\P(X = x)$.

::: {.panel-tabset}

#### Python

```{python}
#| code-fold: false
import pandas as pd

auto = pd.read_csv("data/auto.csv")

by_cyl = auto.groupby("cylinders")["mpg"].agg(["count", "mean"])
by_cyl["P(X = x)"] = by_cyl["count"] / by_cyl["count"].sum()
by_cyl = by_cyl.rename(columns={"mean": "E(Y | X = x)"})
by_cyl = by_cyl[["count", "P(X = x)", "E(Y | X = x)"]]

lie_rhs = (by_cyl["P(X = x)"] * by_cyl["E(Y | X = x)"]).sum()
lie_lhs = auto["mpg"].mean()

by_cyl
```

#### R

```{r}
#| label: lie-auto-r
#| code-fold: false
auto <- read.csv("data/auto.csv")

counts <- table(auto$cylinders)
cond_mean <- tapply(auto$mpg, auto$cylinders, mean)
by_cyl <- data.frame(
  count = as.vector(counts),
  "P(X = x)" = as.vector(counts) / sum(counts),
  "E(Y | X = x)" = as.vector(cond_mean),
  row.names = names(counts),
  check.names = FALSE
)

lie_rhs <- sum(by_cyl[["P(X = x)"]] * by_cyl[["E(Y | X = x)"]])
lie_lhs <- mean(auto$mpg)

by_cyl
```

:::

```{r}
#| include: false
# The claims the prose makes: both languages agree, and the weighted
# average of the conditional means equals the overall mean.
stopifnot(
  abs(lie_lhs - reticulate::py$lie_lhs) < 1e-9,
  abs(lie_rhs - reticulate::py$lie_rhs) < 1e-9,
  abs(lie_lhs - lie_rhs) < 1e-9,
  by_cyl[["E(Y | X = x)"]][1] < by_cyl[["E(Y | X = x)"]][2],
  by_cyl[["E(Y | X = x)"]][5] < by_cyl[["E(Y | X = x)"]][4]
)
```

The conditional mean $\E{Y \mid X = x}$ is the function $g(x)$ of
@def-cond-expectation-function.
Four-cylinder cars average
`{r} sprintf("%.1f", by_cyl["4", "E(Y | X = x)"])`
miles per gallon, and eight-cylinder cars average
`{r} sprintf("%.1f", by_cyl["8", "E(Y | X = x)"])`.
Weighting the `{r} nrow(by_cyl)` conditional means by $\P(X = x)$ gives
`{r} round(lie_rhs, 3)`,
which equals the overall average miles per gallon,
`{r} round(lie_lhs, 3)`,
as @thm-lie states: $\E{\E{Y \mid X}} = \E{Y}$.

:::
