:::{#exr-fubini-joint-mixed}

#### Mixed case: one continuous variable, one discrete variable

Let $Y \sim \mathrm{Bernoulli}(0.6)$ and,
given $Y = y$, let $X \mid Y = y \sim \mathrm{Uniform}(0,\, y + 1)$.

Compute $\E{X}$.

:::

::::{.solution}

Compute $\E{X}$ using @cor-fubini-joint with $\measure_X = \text{Lebesgue measure}$,
$\measure_Y = \text{counting measure}$, and $h(x, y) = x$.

The joint density w.r.t. Lebesgue $\times$ counting measure is
$f_{X,Y}(x, y) = f_{X \mid Y}(x \mid y)\,\P(Y = y)$:

$$
\ba
f_{X,Y}(x,\, 0) &= 1 \cdot 0.4 = 0.4 &&\text{ for } x \in [0,1];\\
f_{X,Y}(x,\, 1) &= \tfrac{1}{2} \cdot 0.6 = 0.3 &&\text{ for } x \in [0,2].
\ea
$$

By @cor-fubini-joint (mixed case):

$$
\ba
\E{X}
&= \sum_{y \in \{0,1\}} \int_0^{y+1} x\,f_{X,Y}(x,\, y)\,dx
&& \text{(joint-distribution form of Fubini--Tonelli, mixed case)} \\
&= \int_0^1 x \cdot 0.4\,dx
   + \int_0^2 x \cdot 0.3\,dx
&& \text{(expand the sum over } y \in \{0, 1\} \text{; substitute } f_{X,Y} \text{)} \\
&= 0.4 \cdot \frac{1}{2} + 0.3 \cdot 2
&& \text{(} \textstyle\int_0^b x\,dx = b^2/2 \text{)} \\
&= 0.2 + 0.6
&& \text{(multiply)} \\
&= 0.8
&& \text{(add)}
\ea
$$

As a check using the law of iterated expectations (@thm-lie):
$\E{X \mid Y = 0} = \tfrac{1}{2}$ and $\E{X \mid Y = 1} = 1$, so
$\E{X} = \tfrac{1}{2}(0.4) + 1(0.6) = 0.2 + 0.6 = 0.8$.

:::{#fig-fubini-joint-mixed}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()

x_fine <- seq(0, 2, by = 0.005)
df <- data.frame(
  x = c(x_fine[x_fine <= 1], x_fine),
  density = c(rep(0.4, sum(x_fine <= 1)), rep(0.3, length(x_fine))),
  label = c(
    rep("Y = 0  (P = 0.4)", sum(x_fine <= 1)),
    rep("Y = 1  (P = 0.6)", length(x_fine))
  )
)

plotly::plot_ly(
  df, x = ~x, y = ~density, color = ~label,
  colors = c("steelblue", "tomato")
) |>
  plotly::add_lines() |>
  plotly::layout(
    xaxis = list(title = "x"),
    yaxis = list(title = "f<sub>X,Y</sub>(x, y)", range = c(0, 0.55)),
    legend = list(title = list(text = "Y value"))
  )
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis
cat(
  "*(This line plot is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Joint density $f_{X,Y}(x, y) = f_{X \mid Y}(x \mid y)\,\P(Y = y)$
for each value of the discrete variable $Y$.
The area under each component integrates to $\P(Y = y)$:
$0.4 \cdot 1 = 0.4$ (blue) and $0.3 \cdot 2 = 0.6$ (red),
summing to 1.

:::

::::

::: {.remark .notes}

$Y$ takes only finitely many values here (two), so
$\sum_{y \in \{0,1\}} \int_0^{y+1} x\,f_{X,Y}(x,\,y)\,dx$
is just linearity of the integral applied to a two-term sum
--- $\int g + \int k = \int (g + k)$ ---
not a genuine interchange of summation and integration order.
If $Y$ had a countably infinite range instead,
the sum of integrals would be an infinite series,
and @cor-fubini-joint's guarantee would be load-bearing.

:::
