:::{#exr-fubini-joint-disc}

#### Expectation of a sum, given a joint PMF

Let $(X, Y)$ be discrete with joint probability mass function:

| | $Y = 0$ | $Y = 1$ |
|:---:|:---:|:---:|
| $X = 0$ | $0.2$ | $0.3$ |
| $X = 1$ | $0.1$ | $0.4$ |

Compute $\E{X + Y}$.

:::

::::{.solution}

Treating $(X, Y)$ as a single discrete random object
taking one of the four values $(0,0)$, $(0,1)$, $(1,0)$, $(1,1)$,
LOTUS (@thm-lotus) applied to $g(x, y) = x + y$ gives directly:

$$
\ba
\E{X + Y}
&= \sum_{x \in \{0,1\}} \sum_{y \in \{0,1\}} (x + y)\,\P(X = x,\, Y = y) && \text{(LOTUS)} \\
&= (0{+}0)(0.2) + (0{+}1)(0.3) + (1{+}0)(0.1) + (1{+}1)(0.4) && \text{(expand the four terms)} \\
&= 0 + 0.3 + 0.1 + 0.8 && \text{(multiply)} \\
&= 1.2 && \text{(add)}
\ea
$$

As a check: $\E{X} = 0(0.5) + 1(0.5) = 0.5$ and
$\E{Y} = 0(0.3) + 1(0.7) = 0.7$,
so $\E{X + Y} = \E{X} + \E{Y} = 1.2$
by [linearity of expectation](#thm-linearity-expectation).

:::{#fig-fubini-joint-disc}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()

x_labs <- c("X=0", "X=0", "X=1", "X=1")
y_labs <- c("Y=0", "Y=1", "Y=0", "Y=1")
probs <- c(0.2, 0.3, 0.1, 0.4)

plotly::plot_ly(
  x = ~y_labs, y = ~probs, color = ~x_labs,
  colors = c("steelblue", "tomato"),
  type = "bar"
) |>
  plotly::layout(
    barmode = "group",
    xaxis = list(title = "Y"),
    yaxis = list(title = "P(X = x, Y = y)", range = c(0, 0.5)),
    legend = list(title = list(text = "X value"))
  )
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis
cat(
  "*(This bar chart is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Joint probability mass function $\P(X = x, Y = y)$.
Marginal totals: $\P(X = 0) = 0.5$, $\P(X = 1) = 0.5$,
$\P(Y = 0) = 0.3$, $\P(Y = 1) = 0.7$.

:::

::::

::: {.remark .notes}

Note that $X$ and $Y$ are **not** independent here:
$\P(X = 0, Y = 0) = 0.2 \neq 0.15 = \P(X = 0)\,\P(Y = 0)$.
LOTUS applies regardless, since it requires only the
*actual* joint mass function, not independence.

There are only four $(x,y)$ pairs here,
so summing them in any order --- row by row, column by column,
or any other listing --- gives the same total
by ordinary commutativity and associativity of addition;
no result about interchanging summation order is needed.
Once the support is countably infinite,
exchanging the order of summation needs a justification,
which the Fubini--Tonelli theorem provides.

:::
