:::{#thm-fubini-tonelli-measure}

#### Fubini--Tonelli theorem, measure-theoretic form

Let $\measure_1$ and $\measure_2$ be $\sigma$-finite [measures](https://morrison-lab.github.io/mds/measures.html#def-measure)
on $\sigma$-algebras of subsets of sets $S_1$ and $S_2$,
and let $f$ be a measurable function on $S_1 \times S_2$.
If either

(a) $f \ge 0$ (Tonelli), or

(b) $\int_{S_1 \times S_2} \abs{f}\,d(\measure_1 \otimes \measure_2) < \infty$ (Fubini),

then the integral of $f$ against the product measure $\measure_1 \otimes \measure_2$
equals both iterated integrals:

$$
\ba
\int_{S_1 \times S_2} f\,d(\measure_1 \otimes \measure_2)
&= \int_{S_1}\paren{\int_{S_2} f(s_1, s_2)\,d\measure_2(s_2)}\,d\measure_1(s_1)
\\&= \int_{S_2}\paren{\int_{S_1} f(s_1, s_2)\,d\measure_1(s_1)}\,d\measure_2(s_2).
\ea
$$

In case (b), each inner integral is finite
except on a set of measure $0$, which does not affect the outer integral.

:::

::: remark

For expectations, we use this measure-theoretic form of the
[Fubini--Tonelli theorem](https://morrison-lab.github.io/mds/calculus.html#thm-fubini-tonelli),
which lets us exchange the order of integration (or summation)
over a product of $\sigma$-finite measure spaces,
provided the integrand is non-negative (Tonelli)
or absolutely integrable (Fubini).
Its proof is beyond these notes' scope
[@billingsley1995probability, Theorem 18.3].
Lebesgue measure on the real line and
[counting measure](https://morrison-lab.github.io/mds/measures.html#def-counting-measure) on a countable set
are both $\sigma$-finite,
which gives the theorem a form stated in terms of a joint distribution.

:::

{{< slidebreak >}}

:::{#cor-fubini-joint}

#### Joint-distribution form (without independence; corollary of Fubini--Tonelli)

Let $(X, Y)$ be jointly distributed random variables
whose joint distribution has a density $f_{X,Y}$
with respect to a product of $\sigma$-finite reference measures
$\measure_X \otimes \measure_Y$ on $\rangef{X} \times \rangef{Y}$,
and let $h : \rangef{X} \times \rangef{Y} \to \reals$ be measurable.
If either

(a) $h(X, Y) \ge 0$ almost surely, or

(b) $\E{\abs{h(X, Y)}} < \infty$,

then the expectation of $h(X, Y)$ can be written as an iterated integral
against $f_{X,Y}$, with the order of integration exchangeable:

$$
\ba
\E{h(X, Y)}
&= \int_{\rangef{X}}\paren{\int_{\rangef{Y}} h(x, y)\,f_{X,Y}(x, y)\,d\measure_Y(y)}\,d\measure_X(x)
\\&= \int_{\rangef{Y}}\paren{\int_{\rangef{X}} h(x, y)\,f_{X,Y}(x, y)\,d\measure_X(x)}\,d\measure_Y(y).
\ea
$$

The choice of reference measures covers three cases:

- **Both continuous:**

  $$
  \begin{aligned}
  \measure_X &= \measure_Y \\
  &= \text{Lebesgue measure};
  \end{aligned}
  $$

  $f_{X,Y}$ is the joint probability density function (PDF),
  and $\int_{\rangef{X}} g(x)\,d\measure_X(x) = \int_{\rangef{X}} g(x)\,dx$.
- **Both discrete:**

  $$
  \begin{aligned}
  \measure_X &= \measure_Y \\
  &= \text{counting measure};
  \end{aligned}
  $$

  $f_{X,Y}(x,y) = \P(X = x,\, Y = y)$ is the joint probability mass function (PMF),
  and $\int_{\rangef{X}} g(x)\,d\measure_X(x) = \sum_{x \in \rangef{X}} g(x)$.
- **Mixed** (one continuous, one discrete):
  one reference measure is Lebesgue and the other is counting;
  $f_{X,Y}$ is the [joint density-mass function](random-variables.qmd#def-joint-density-mass),
  $f_{X,Y}(x,y) = f_{X \mid Y}(x \mid y)\,\P(Y = y)$
  (or $\P(X = x \mid Y = y)\,f_Y(y)$ if $X$ is discrete and $Y$ continuous),
  and the iterated integrals combine an ordinary integral with a sum.
  The conditional densities/PMFs here are defined the same way as in
  @def-cond-mixed, just conditioning on $Y$ instead of $X$.

:::

::: proof
Apply @thm-fubini-tonelli-measure with
$\measure_1 = \measure_X$ and $\measure_2 = \measure_Y$ to the integrand
$h(x,y)\,f_{X,Y}(x,y)$ on $\rangef{X} \times \rangef{Y}$.
Lebesgue measure and counting measure on a countable set are each
$\sigma$-finite, so $\measure_X \otimes \measure_Y$ is $\sigma$-finite in all three cases.
The relevant condition is (a) when $h \ge 0$ and
(b) when $\E{\abs{h(X, Y)}} < \infty$.
Independence is not required.
When $X$ and $Y$ are independent,
$f_{X,Y}(x,y) = f_X(x)\,f_Y(y)$
(or $\P(X=x,Y=y) = \P(X=x)\,\P(Y=y)$ in the discrete case),
and the iterated integrals factor into separate integrals over the marginals.
:::

{{< slidebreak >}}

:::{#exm-fubini-prob}

#### Expectation of a product of independent variables

Let $X \sim \mathrm{Uniform}(0, 1)$ and $Y \sim \mathrm{Uniform}(0, 2)$,
independently distributed.
Compute $\E{XY}$.

We apply @cor-fubini-joint (both-continuous case) with $h(x, y) = xy$.
Since $X$ and $Y$ are independent with densities $f_X(x) = 1$ on $[0,1]$
and $f_Y(y) = \tfrac{1}{2}$ on $[0,2]$,
the joint density factors as

$$
\begin{aligned}
f_{X,Y}(x,y) &= f_X(x)\,f_Y(y) \\
&= \tfrac{1}{2},
\end{aligned}
$$

and

$$
\begin{aligned}
\measure_X &= \measure_Y \\
&= \text{Lebesgue measure}:
\end{aligned}
$$

$$
\ba
\E{XY}
&= \int_0^1 \paren{\int_0^2 xy \cd \tfrac{1}{2}\,dy}\,dx
&& \text{(joint-distribution form of Fubini--Tonelli)}
\\&= \int_0^1 x\paren{\frac{1}{2}\int_0^2 y\,dy}\,dx
&& \text{(factor constants out of the inner integral)}
\\&= \int_0^1 x \cd \frac{1}{2} \cd \sb{\frac{y^2}{2}}_0^2\,dx
&& \text{(antiderivative of } y \text{)}
\\&= \int_0^1 x \cd \frac{1}{2} \cd 2\,dx
&& \text{(evaluate at the bounds)}
\\&= \int_0^1 x\,dx
&& \text{(simplify)}
\\&= \frac{1}{2}
&& \text{(integrate)}
\ea
$$

As a check: $\E{X} = \tfrac{1}{2}$, $\E{Y} = 1$, and $\E{X}\E{Y} = \tfrac{1}{2}$.

:::

{{< slidebreak >}}

::::{#exm-fubini-prob-fail}

#### When independence fails: a counterexample

Correctly applying @cor-fubini-joint requires the *actual* joint density $f_{X,Y}$
--- not the product of marginals $f_X(x)\,f_Y(y)$,
which is valid only when $X$ and $Y$ are independent.
Using the wrong joint density gives the wrong answer.

Let $X \sim \mathrm{Uniform}(0, 1)$ and set $Y = X$
(so $X$ and $Y$ are perfectly correlated and **not** independent).

**True expectation:**

$$
\ba
\E{XY}
&= \E{X \cd X}
&& \text{(} Y = X \text{)}
\\&= \E{X^2}
&& \text{(} X \cd X = X^2 \text{)}
\\&= \int_0^1 x^2\,dx
&& \text{(LOTUS, with the uniform density } 1 \text{ on } [0, 1] \text{)}
\\&= \frac{1}{3}
&& \text{(integrate)}
\ea
$$

**Erroneously applying the product-measure formula:**

Note that Fubini--Tonelli's own conditions still hold here ($h(x,y) = xy$
is nonnegative and integrable), so the error is not a failure of
Fubini--Tonelli itself.
Rather, the error is using the *wrong measure*: the joint distribution
of $(X, X)$ is concentrated on the diagonal
$\{(x, x) : x \in [0, 1]\} \subset [0, 1]^2$,
which has Lebesgue measure zero in $\reals^2$.
The joint distribution is therefore **not** absolutely continuous with
respect to two-dimensional Lebesgue measure, so **no joint density
$f_{X,Y}$ on $[0, 1]^2$ exists**, which is the reference density
@cor-fubini-joint requires.

The following calculation is what someone would *erroneously* write
if they assumed independence and used $f_X(x)\,f_Y(y)$ as a "joint density"
--- a function that does not in fact correspond to the joint distribution of $(X, X)$.
The marginals $X \sim \mathrm{Uniform}(0,1)$ and $Y \sim \mathrm{Uniform}(0,1)$
do have densities

$$
\begin{aligned}
f_X &= f_Y \\
&= 1,
\end{aligned}
$$

but the *product* $f_X(x)\,f_Y(y) = 1$ on $[0, 1]^2$
is the density of an *independent* pair, not of $(X, X)$:

$$
\ba
\int_0^1\!\int_0^1 xy \cd f_X(x) \cd f_Y(y)\,dy\,dx
&= \int_0^1\!\int_0^1 xy\,dy\,dx
&& \text{(} f_X(x)\,f_Y(y) = 1 \text{ on } [0, 1]^2 \text{)}
\\&= \int_0^1 x\paren{\int_0^1 y\,dy}\,dx
&& \text{(factor } x \text{ out of the inner integral)}
\\&= \int_0^1 x \cd \frac{1}{2}\,dx
&& \text{(} \textstyle\int_0^1 y\,dy = \tfrac{1}{2} \text{)}
\\&= \frac{1}{4}
&& \text{(} \textstyle\int_0^1 \tfrac{x}{2}\,dx = \tfrac{1}{4} \text{)}
\ea
$$

This calculation recovers $\E{XY}$ for *independent* uniforms ($\tfrac{1}{4}$),
not $\E{XX}$ for the perfectly correlated pair ($\tfrac{1}{3}$).
The lesson is that @cor-fubini-joint requires the *actual* joint density $f_{X,Y}$.
For independent $(X, Y)$, this density factors as $f_X(x)\,f_Y(y)$;
for dependent $(X, Y)$, $f_{X,Y}$ need not factor
--- and for $(X, X)$, no joint density on $\reals^2$ exists at all,
so @cor-fubini-joint simply does not apply.

:::{#fig-fubini-prob-fail}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()
set.seed(204)
n <- 400
x_dep <- runif(n)
y_dep <- x_dep
x_ind <- runif(n)
y_ind <- runif(n)

plotly::plot_ly() |>
  plotly::add_trace(
    type = "scatter", mode = "markers",
    x = x_ind, y = y_ind,
    name = "Assumed independent (X<sub>1</sub>, X<sub>2</sub>)",
    marker = list(size = 5, color = "#999999", opacity = 0.5)
  ) |>
  plotly::add_trace(
    type = "scatter", mode = "markers",
    x = x_dep, y = y_dep,
    name = "Actual (X, X) on diagonal",
    marker = list(size = 6, color = "#b40426")
  ) |>
  plotly::layout(
    xaxis = list(title = "x", range = c(0, 1), scaleanchor = "y"),
    yaxis = list(title = "y", range = c(0, 1)),
    legend = list(orientation = "h", y = -0.2)
  )
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis

cat(
  "*(This scatter plot is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Samples from the joint distribution of $(X, X)$ (red, on the diagonal)
versus an independent pair $(X_1, X_2)$ with the same marginals
(grey, scattered over $[0, 1]^2$).
The actual joint mass for $(X, X)$ is concentrated on a 1-dimensional diagonal
--- a set of Lebesgue measure zero in $\reals^2$ ---
so no joint density on $[0, 1]^2$ exists,
and the "$f_X(x)\,f_Y(y) = 1$" calculation integrates against the wrong measure
(the grey distribution).

:::

::::

{{< slidebreak >}}

::::{#exm-fubini-joint}

#### Both-continuous case: joint PDF on a non-rectangular support

Let $(X, Y)$ have joint density
$f_{X,Y}(x, y) = 2$ for $0 \le x \le y \le 1$
(and $0$ otherwise).
Compute $\E{X + Y}$.

By @cor-fubini-joint:

$$
\ba
\E{X + Y}
&= \int_0^1\!\int_0^y (x + y) \cd 2\,dx\,dy
&& \text{(joint-distribution form of Fubini--Tonelli, over the support } 0 \le x \le y \le 1 \text{)}
\\&= 2\int_0^1 \sb{\frac{x^2}{2} + xy}_{x=0}^{x=y}\,dy
&& \text{(antiderivative in } x \text{)}
\\&= 2\int_0^1 \paren{\frac{y^2}{2} + y^2}\,dy
&& \text{(evaluate at the bounds)}
\\&= 2\int_0^1 \frac{3y^2}{2}\,dy
&& \text{(add)}
\\&= 3\int_0^1 y^2\,dy
&& \text{(simplify the constant)}
\\&= 3 \cd \frac{1}{3}
&& \text{(integrate)}
\\&= 1
&& \text{(multiply)}
\ea
$$

:::{#fig-fubini-joint}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()
n_grid <- 51
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) {
  z <- rep(2, length(x))
  z[x > y] <- NA
  z
})

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "f(x, y)", range = c(0, 2.5)),
    camera = list(eye = list(x = 1.6, y = -1.6, z = 0.8))
  ))
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis

cat(
  "*(This 3D surface is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Joint density $f_{X,Y}(x, y) = 2$ on the triangular support
$\{(x, y) : 0 \le x \le y \le 1\}$, and zero elsewhere.
The total "volume" under the density is $2 \cdot \tfrac{1}{2} = 1$, as required.

:::

::::

{{< slidebreak >}}

:::{#thm-lie}
#### Law of iterated expectations

For any two random variables $X$ and $Y$ with $\E{\abs{Y}} < \infty$:

$$\E{Y} = \E{\E{Y \mid X}}$$

:::

{{< slidebreak >}}

::: proof

**Discrete case.**
When $X$ and $Y$ are discrete,
applying @def-expectation to $\E{\E{Y \mid X}}$
and then the [law of total probability](probability-basics.qmd#thm-total-prob)
applied to the countable partition $\{X = x : x \in \rangef{X}\}$:

$$
\ba
\E{\E{Y \mid X}}
&= \sum_{x \in \rangef{X}} \E{Y \mid X=x} \cd \P(X=x) && \text{(LOTUS for the function } x \mapsto \E{Y \mid X=x} \text{)}
\\&= \sum_{x \in \rangef{X}} \paren{\sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x)} \cd \P(X=x) && \text{(definition of conditional expectation } \E{Y \mid X=x} \text{)}
\\&= \sum_{y \in \rangef{Y}} y \cd \sum_{x \in \rangef{X}} \P(Y=y \mid X=x) \cd \P(X=x) && \text{(exchange order of summation: Fubini, since } \E{\abs{Y}} < \infty \text{)}
\\&= \sum_{y \in \rangef{Y}} y \cd \P(Y=y) && \text{(law of total probability over } X \text{)}
\\&= \E{Y} && \text{(definition of expectation of } Y \text{)}
\ea
$$

**Continuous case.**
When $X$ and $Y$ are continuous:

$$
\ba
\E{\E{Y \mid X}}
&= \int_{x \in \rangef{X}} \E{Y \mid X=x} \cd \p(X=x)\, dx
&& \text{(LOTUS)}
\\&= \int_{x \in \rangef{X}} \paren{\int_{y \in \rangef{Y}} y \cd \p(Y=y \mid X=x)\, dy} \cd \p(X=x)\, dx
&& \text{(definition of conditional expectation)}
\\&= \int_{x \in \rangef{X}} \int_{y \in \rangef{Y}} y \cd \p(X=x, Y=y)\, dy\, dx
&& \text{(definition of conditional density: } \p(Y=y \mid X=x)\,\p(X=x) = \p(X=x, Y=y) \text{)}
\\&= \int_{y \in \rangef{Y}} y \cd \paren{\int_{x \in \rangef{X}} \p(X=x, Y=y)\, dx}\, dy
&& \text{(Fubini's theorem, since } \E{\abs{Y}} < \infty \text{)}
\\&= \int_{y \in \rangef{Y}} y \cd \p(Y=y)\, dy
&& \text{(marginal density from a joint density)}
\\&= \E{Y}
&& \text{(definition of expectation)}
\ea
$$

**Mixed case, $X$ discrete and $Y$ continuous.**
With the [joint density-mass function](random-variables.qmd#def-joint-density-mass)
$\p(X = x,\, Y = y)$,
@cor-fubini-joint applies with $h(x, y) = y$,
counting measure for $X$, and Lebesgue measure for $Y$;
its condition (b) holds because $\E{\abs{Y}} < \infty$.
The sum runs over the $x$ with $\P(X = x) > 0$,
where @def-cond-mixed defines $\E{Y \mid X = x}$:

$$
\ba
\E{\E{Y \mid X}}
&= \sum_{x \in \rangef{X}} \E{Y \mid X=x} \cd \P(X=x)
&& \text{(LOTUS, discrete case)}
\\&= \sum_{x \in \rangef{X}} \paren{\int_{\rangef{Y}} y \cd \p(Y=y \mid X=x)\,dy} \cd \P(X=x)
&& \text{(definition of conditional expectation, mixed case)}
\\&= \sum_{x \in \rangef{X}} \int_{\rangef{Y}} y \cd \p(Y=y \mid X=x) \cd \P(X=x)\,dy
&& \text{(move the constant } \P(X=x) \text{ inside the integral)}
\\&= \sum_{x \in \rangef{X}} \int_{\rangef{Y}} y \cd \p(X=x,\, Y=y)\,dy
&& \text{(definition of the conditional density, mixed case)}
\\&= \E{Y}
&& \text{(joint-distribution form of Fubini--Tonelli, with } h(x, y) = y \text{)}
\ea
$$

**Mixed case, $X$ continuous and $Y$ discrete.**
The same steps apply with the integral and the sum swapped,
over the $x$ with $\p(X = x) > 0$;
at any other $x$, $\p(X = x,\, Y = y) = 0$ for every $y$,
because these non-negative terms sum to $\p(X = x)$:

$$
\ba
\E{\E{Y \mid X}}
&= \int_{\rangef{X}} \E{Y \mid X=x} \cd \p(X=x)\,dx
&& \text{(LOTUS, continuous case)}
\\&= \int_{\rangef{X}} \paren{\sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x)} \cd \p(X=x)\,dx
&& \text{(definition of conditional expectation, mixed case)}
\\&= \int_{\rangef{X}} \sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x) \cd \p(X=x)\,dx
&& \text{(move the constant } \p(X=x) \text{ inside the sum)}
\\&= \int_{\rangef{X}} \sum_{y \in \rangef{Y}} y \cd \p(X=x,\, Y=y)\,dx
&& \text{(definition of the conditional PMF, mixed case)}
\\&= \E{Y}
&& \text{(joint-distribution form of Fubini--Tonelli, with } h(x, y) = y \text{)}
\ea
$$

:::

::: {.remark .notes}

Alternate names for this identity include:
the **tower rule**,
the **tower property**,
the **law of total expectation**,
and the **smoothing theorem**.

:::

```{r}
#| include: false
reticulate::py_require(c("numpy", "pandas"))
```

{{< include pds/_subfiles/_exm-lie-auto.qmd >}}

{{< slidebreak >}}

:::{#thm-conditional-lie}
#### Conditional law of iterated expectations

For random variables $X$, $Y$, and $Z$ with $\E{\abs{Y}} < \infty$:

$$\E{Y \mid Z} = \E{\E{Y \mid X,Z} \mid Z}$$

:::

{{< slidebreak >}}

::: proof

Fix a value $z$ with positive probability (discrete case) or positive density (continuous case).
Conditioning every probability on $Z = z$ gives a probability measure,
and the proof of @thm-lie goes through under it, line by line.

**Discrete case.**

$$
\ba
\E{\E{Y \mid X,Z} \mid Z=z}
&= \sum_{x \in \rangef{X}} \E{Y \mid X=x,Z=z} \cd \P(X=x \mid Z=z) && \text{(expectation given } Z=z \text{)}
\\&= \sum_{x \in \rangef{X}} \paren{\sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x,Z=z)} \cd \P(X=x \mid Z=z) && \text{(definition of } \E{Y \mid X=x,Z=z} \text{)}
\\&= \sum_{y \in \rangef{Y}} y \cd \sum_{x \in \rangef{X}} \P(Y=y \mid X=x,Z=z) \cd \P(X=x \mid Z=z) && \text{(exchange order of summation)}
\\&= \sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid Z=z) && \text{(law of total probability given } Z=z \text{)}
\\&= \E{Y \mid Z=z} && \text{(definition of conditional expectation given } Z=z \text{)}
\ea
$$

**Continuous case.**

$$
\ba
\E{\E{Y \mid X,Z} \mid Z=z}
&= \int_{x \in \rangef{X}} \E{Y \mid X=x,Z=z} \cd \p(X=x \mid Z=z)\, dx
&& \text{(expectation given } Z=z \text{)}
\\&= \int_{x \in \rangef{X}} \paren{\int_{y \in \rangef{Y}} y \cd \p(Y=y \mid X=x,Z=z)\, dy} \cd \p(X=x \mid Z=z)\, dx
&& \text{(definition of } \E{Y \mid X=x,Z=z} \text{)}
\\&= \int_{x \in \rangef{X}} \int_{y \in \rangef{Y}} y \cd \p(X=x, Y=y \mid Z=z)\, dy\, dx
&& \text{(product of conditional densities is the joint conditional density)}
\\&= \int_{y \in \rangef{Y}} y \cd \paren{\int_{x \in \rangef{X}} \p(X=x, Y=y \mid Z=z)\, dx}\, dy
&& \text{(Fubini's theorem)}
\\&= \int_{y \in \rangef{Y}} y \cd \p(Y=y \mid Z=z)\, dy
&& \text{(marginalize over } x \text{)}
\\&= \E{Y \mid Z=z}
&& \text{(definition of conditional expectation given } Z=z \text{)}
\ea
$$

Since the two sides agree at every such $z$,
they agree as random variables (functions of $Z$):
$\E{Y \mid Z} = \E{\E{Y \mid X,Z} \mid Z}$.

:::

::: {.remark .notes}

This identity is the tower rule
applied conditionally on $Z$.

:::

{{< slidebreak >}}

:::{#exm-lie}

#### Marginal expectation from conditional expectations

Suppose $X$ is a binary random variable indicating treatment assignment ($X=1$ treated, $X=0$ control),
with $\P(X=1) = 0.5$,
and suppose the outcome $Y$ has conditional expectations:

$$\E{Y \mid X=1} = 10, \quad \E{Y \mid X=0} = 6$$

By the law of iterated expectations (@thm-lie):

$$
\ba
\E{Y}
&= \E{\E{Y \mid X}}
&& \text{(law of iterated expectations)}
\\&= \E{Y \mid X=1} \cd \P(X=1) + \E{Y \mid X=0} \cd \P(X=0)
&& \text{(LOTUS over the two values of } X \text{)}
\\&= 10 \cd 0.5 + 6 \cd 0.5
&& \text{(substitute)}
\\&= 5 + 3
&& \text{(multiply)}
\\&= 8
&& \text{(add)}
\ea
$$

:::

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-disc-infinite.qmd >}}

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-mixed.qmd >}}

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-mixed-infinite.qmd >}}
