For expectations against probability measures we use the
measure-theoretic form of Fubini--Tonelli,
which holds on any $\sigma$-finite measure space.
The $\sigma$-finiteness condition is automatic for probability measures
(every probability measure is finite, hence $\sigma$-finite),
yielding @cor-fubini-joint directly.

{{< slidebreak >}}

:::{#cor-fubini-joint}

#### Joint-distribution form (without independence; corollary of Fubini--Tonelli)

Let $(X, Y)$ be jointly distributed random variables
whose joint distribution has a density $f_{X,Y}$
with respect to a product of $\sigma$-finite reference measures
$\mu_X \otimes \mu_Y$ on $\rangef{X} \times \rangef{Y}$,
and let $h : \rangef{X} \times \rangef{Y} \to \mathbb{R}$ be measurable.
If either

(a) $h(X, Y) \ge 0$ almost surely, or

(b) $\E{\abs{h(X, Y)}} < \infty$,

then the expectation of $h(X, Y)$ can be written as an iterated integral
against $f_{X,Y}$, with the order of integration exchangeable:

$$
\ba
\E{h(X, Y)}
&= \int_{\rangef{X}}\paren{\int_{\rangef{Y}} h(x, y)\,f_{X,Y}(x, y)\,d\mu_Y(y)}\,d\mu_X(x)
\\&= \int_{\rangef{Y}}\paren{\int_{\rangef{X}} h(x, y)\,f_{X,Y}(x, y)\,d\mu_X(x)}\,d\mu_Y(y).
\ea
$$

The choice of reference measures covers three cases:

- **Both continuous:** $\mu_X = \mu_Y = \text{Lebesgue measure}$;
  $f_{X,Y}$ is the joint probability density function (PDF),
  and $\int g(x)\,d\mu_X(x) = \int g(x)\,dx$.
- **Both discrete:** $\mu_X = \mu_Y = \text{counting measure}$;
  $f_{X,Y}(x,y) = \P(X = x,\, Y = y)$ is the joint probability mass function (PMF),
  and $\int g(x)\,d\mu_X(x) = \sum_{x \in \rangef{X}} g(x)$.
- **Mixed** (one continuous, one discrete):
  one reference measure is Lebesgue and the other is counting;
  $f_{X,Y}(x,y) = f_{X \mid Y}(x \mid y)\,\P(Y = y)$
  (or $\P(X = x \mid Y = y)\,f_Y(y)$ if $X$ is discrete and $Y$ continuous),
  and the iterated integrals combine an ordinary integral with a sum.
  The conditional densities/PMFs here are defined the same way as in
  @def-cond-mixed, just conditioning on $Y$ instead of $X$.

:::

::: proof
Apply Fubini--Tonelli with
$\mu_1 = \mu_X$ and $\mu_2 = \mu_Y$ to the integrand
$h(x,y)\,f_{X,Y}(x,y)$ on $\rangef{X} \times \rangef{Y}$.
Lebesgue measure and counting measure on a countable set are each
$\sigma$-finite, so $\mu_X \otimes \mu_Y$ is $\sigma$-finite in all three cases.
The relevant condition is (a) when $h \ge 0$ and
(b) when $\E{\abs{h(X, Y)}} < \infty$.
Independence is not required.
When $X$ and $Y$ are independent,
$f_{X,Y}(x,y) = f_X(x)\,f_Y(y)$
(or $\P(X=x,Y=y) = \P(X=x)\,\P(Y=y)$ in the discrete case),
and the iterated integrals factor into separate integrals over the marginals.
:::

{{< slidebreak >}}

:::{#exm-fubini-prob}

#### Expectation of a product of independent variables

Let $X \sim \mathrm{Uniform}(0, 1)$ and $Y \sim \mathrm{Uniform}(0, 2)$,
independently distributed.
Compute $\E{XY}$.

We apply @cor-fubini-joint (both-continuous case) with $h(x, y) = xy$.
Since $X$ and $Y$ are independent with densities $f_X(x) = 1$ on $[0,1]$
and $f_Y(y) = \tfrac{1}{2}$ on $[0,2]$,
the joint density factors as $f_{X,Y}(x,y) = f_X(x)\,f_Y(y) = \tfrac{1}{2}$,
and $\mu_X = \mu_Y = \text{Lebesgue measure}$:

$$
\ba
\E{XY}
&= \int_0^1 \paren{\int_0^2 xy \cd \tfrac{1}{2}\,dy}\,dx
\\&= \int_0^1 x\paren{\frac{1}{2}\int_0^2 y\,dy}\,dx
\\&= \int_0^1 x \cd \frac{1}{2} \cd \sb{\frac{y^2}{2}}_0^2\,dx
\\&= \int_0^1 x \cd \frac{1}{2} \cd 2\,dx
\\&= \int_0^1 x\,dx
\\&= \frac{1}{2}
\ea
$$

As a check: $\E{X} = \tfrac{1}{2}$, $\E{Y} = 1$, and $\E{X}\E{Y} = \tfrac{1}{2}$.

:::

{{< slidebreak >}}

::::{#exm-fubini-prob-fail}

#### When independence fails: a counterexample

Correctly applying @cor-fubini-joint requires the *actual* joint density $f_{X,Y}$
--- not the product of marginals $f_X(x)\,f_Y(y)$, which is valid only when $X$ and $Y$
are independent. Using the wrong joint density gives the wrong answer.

Let $X \sim \mathrm{Uniform}(0, 1)$ and set $Y = X$
(so $X$ and $Y$ are perfectly correlated and **not** independent).

**True expectation:**

$$
\E{XY} = \E{X \cd X} = \E{X^2} = \int_0^1 x^2\,dx = \frac{1}{3}
$$

**Erroneously applying the product-measure formula:**

Note that Fubini--Tonelli's own conditions still hold here ($h(x,y) = xy$
is nonnegative and integrable), so the error is not a failure of
Fubini--Tonelli itself.
Rather, the error is using the *wrong measure*: the joint distribution
of $(X, X)$ is concentrated on the diagonal
$\{(x, x) : x \in [0, 1]\} \subset [0, 1]^2$,
which has Lebesgue measure zero in $\mathbb{R}^2$.
The joint distribution is therefore **not** absolutely continuous with
respect to two-dimensional Lebesgue measure, so **no joint density
$f_{X,Y}$ on $[0, 1]^2$ exists**, which is the reference density
@cor-fubini-joint requires.

The following calculation is what someone would *erroneously* write if
they assumed independence and used $f_X(x)\,f_Y(y)$ as a "joint
density" --- a function that does not in fact correspond to the joint
distribution of $(X, X)$. The marginals
$X \sim \mathrm{Uniform}(0,1)$ and $Y \sim \mathrm{Uniform}(0,1)$
do have densities $f_X = f_Y = 1$, but the *product*
$f_X(x)\,f_Y(y) = 1$ on $[0, 1]^2$ is the density of an *independent*
pair, not of $(X, X)$:

$$
\ba
\int_0^1\!\int_0^1 xy \cd f_X(x) \cd f_Y(y)\,dy\,dx
&= \int_0^1\!\int_0^1 xy\,dy\,dx
\\&= \int_0^1 x\paren{\int_0^1 y\,dy}\,dx
\\&= \int_0^1 x \cd \frac{1}{2}\,dx
\\&= \frac{1}{4}
\ea
$$

This calculation recovers $\E{XY}$ for *independent* uniforms ($\tfrac{1}{4}$),
not $\E{XX}$ for the perfectly correlated pair ($\tfrac{1}{3}$).
The lesson is that @cor-fubini-joint requires the *actual* joint density
$f_{X,Y}$. For independent $(X, Y)$, this factors as $f_X(x)\,f_Y(y)$;
for dependent $(X, Y)$, $f_{X,Y}$ need not factor --- and for $(X, X)$,
no joint density on $\mathbb{R}^2$ exists at all, so @cor-fubini-joint
simply does not apply.

:::{#fig-fubini-prob-fail}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()
set.seed(204)
n <- 400
x_dep <- runif(n)
y_dep <- x_dep
x_ind <- runif(n)
y_ind <- runif(n)

plotly::plot_ly() |>
  plotly::add_trace(
    type = "scatter", mode = "markers",
    x = x_ind, y = y_ind,
    name = "Assumed independent (X<sub>1</sub>, X<sub>2</sub>)",
    marker = list(size = 5, color = "#999999", opacity = 0.5)
  ) |>
  plotly::add_trace(
    type = "scatter", mode = "markers",
    x = x_dep, y = y_dep,
    name = "Actual (X, X) on diagonal",
    marker = list(size = 6, color = "#b40426")
  ) |>
  plotly::layout(
    xaxis = list(title = "x", range = c(0, 1), scaleanchor = "y"),
    yaxis = list(title = "y", range = c(0, 1)),
    legend = list(orientation = "h", y = -0.2)
  )
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis

cat(
  "*(This scatter plot is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Samples from the joint distribution of $(X, X)$ (red, on the diagonal)
versus an independent pair $(X_1, X_2)$ with the same marginals (grey,
scattered over $[0, 1]^2$). The actual joint mass for $(X, X)$ is
concentrated on a 1-dimensional diagonal --- a set of Lebesgue measure
zero in $\mathbb{R}^2$ --- so no joint density on $[0, 1]^2$ exists, and
the "$f_X(x)\,f_Y(y) = 1$" calculation integrates against the wrong
measure (the grey distribution).

:::

::::

{{< slidebreak >}}

::::{#exm-fubini-joint}

#### Both-continuous case: joint PDF on a non-rectangular support

Let $(X, Y)$ have joint density
$f_{X,Y}(x, y) = 2$ for $0 \le x \le y \le 1$
(and $0$ otherwise).
Compute $\E{X + Y}$.

By @cor-fubini-joint:

$$
\ba
\E{X + Y}
&= \int_0^1\!\int_0^y (x + y) \cd 2\,dx\,dy
\\&= 2\int_0^1 \sb{\frac{x^2}{2} + xy}_{x=0}^{x=y}\,dy
\\&= 2\int_0^1 \paren{\frac{y^2}{2} + y^2}\,dy
\\&= 2\int_0^1 \frac{3y^2}{2}\,dy
\\&= 3\int_0^1 y^2\,dy
\\&= 3 \cd \frac{1}{3}
\\&= 1
\ea
$$

:::{#fig-fubini-joint}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()
n_grid <- 51
x_seq <- seq(0, 1, length.out = n_grid)
y_seq <- seq(0, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) {
  z <- rep(2, length(x))
  z[x > y] <- NA
  z
})

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(showscale = FALSE) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "f(x, y)", range = c(0, 2.5)),
    camera = list(eye = list(x = 1.6, y = -1.6, z = 0.8))
  ))
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis

cat(
  "*(This 3D surface is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Joint density $f_{X,Y}(x, y) = 2$ on the triangular support
$\{(x, y) : 0 \le x \le y \le 1\}$, and zero elsewhere. The total
"volume" under the density is $2 \cdot \tfrac{1}{2} = 1$, as required.

:::

::::

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-disc-infinite.qmd >}}

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-mixed.qmd >}}

{{< slidebreak >}}

{{< include pds/_subfiles/_exr-fubini-joint-mixed-infinite.qmd >}}

{{< slidebreak >}}

:::{#thm-lie}

#### Law of iterated expectations

For any two random variables $X$ and $Y$:

$$\E{Y} = \E{\E{Y \mid X}}$$

::: notes

Alternate names for this identity include:
the **tower rule**,
the **tower property**,
the **law of total expectation**,
and the **smoothing theorem**.

:::

:::

{{< slidebreak >}}

::: proof

**Discrete case.**
When $X$ and $Y$ are discrete,
applying @def-expectation to $\E{\E{Y \mid X}}$
and then the [law of total probability](probability-basics.qmd#thm-total-prob)
applied to the countable partition $\{X = x : x \in \rangef{X}\}$:

$$
\ba
\E{\E{Y \mid X}}
&= \sum_{x \in \rangef{X}} \E{Y \mid X=x} \cd \P(X=x) && \text{(definition of expectation of } \E{Y \mid X} \text{)}
\\&= \sum_{x \in \rangef{X}} \paren{\sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x)} \cd \P(X=x) && \text{(definition of conditional expectation } \E{Y \mid X=x} \text{)}
\\&= \sum_{y \in \rangef{Y}} y \cd \sum_{x \in \rangef{X}} \P(Y=y \mid X=x) \cd \P(X=x) && \text{(exchange order of summation)}
\\&= \sum_{y \in \rangef{Y}} y \cd \P(Y=y) && \text{(law of total probability over } X \text{)}
\\&= \E{Y} && \text{(definition of expectation of } Y \text{)}
\ea
$$

**Continuous case.**
When $X$ and $Y$ are continuous,
applying @def-expectation to $\E{\E{Y \mid X}}$
and then using @def-cond-expectation for $\E{Y \mid X=x}$:

$$
\ba
\E{\E{Y \mid X}}
&= \int_{x \in \rangef{X}} \E{Y \mid X=x} \cd \p(X=x)\, dx
\\&= \int_{x \in \rangef{X}} \paren{\int_{y \in \rangef{Y}} y \cd \p(Y=y \mid X=x)\, dy} \cd \p(X=x)\, dx
\\&= \int_{y \in \rangef{Y}} y \cd \paren{\int_{x \in \rangef{X}} \p(Y=y \mid X=x) \cd \p(X=x)\, dx}\, dy
\\&= \int_{y \in \rangef{Y}} y \cd \p(Y=y)\, dy
\\&= \E{Y}
\ea
$$

where the third equality exchanges the order of integration by
condition (b) of
Fubini--Tonelli (the absolute-integrability
case, **Fubini's theorem**); this theorem requires $\E{\abs{Y}} < \infty$,
which is implicit in $\E{Y}$ being defined,
and the fourth equality uses
$\int_{x} \p(Y=y \mid X=x) \cd \p(X=x)\, dx = \int_{x} \p(X=x, Y=y)\, dx = \p(Y=y)$
(marginalization of the joint density).

:::

{{< slidebreak >}}

:::{#thm-conditional-lie}
#### Conditional law of iterated expectations

For random variables $X$, $Y$, and $Z$:

$$\E{Y \mid Z} = \E{\E{Y \mid X,Z} \mid Z}$$

::: notes

This identity is the tower rule
applied conditionally on $Z$.

:::

:::

{{< slidebreak >}}

::: proof

For each fixed value $z$ with positive probability or density:

**Discrete case.**
Conditioning on $Z=z$,
and applying the law of total probability
to the partition $\{X=x : x \in \rangef{X}\}$
under the conditional distribution given $Z=z$:

$$
\ba
\E{\E{Y \mid X,Z} \mid Z=z}
&= \sum_{x \in \rangef{X}} \E{Y \mid X=x,Z=z} \cd \P(X=x \mid Z=z) && \text{(definition of expectation given } Z=z \text{)}
\\&= \sum_{x \in \rangef{X}} \paren{\sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid X=x,Z=z)} \cd \P(X=x \mid Z=z) && \text{(definition of conditional expectation } \E{Y \mid X=x,Z=z} \text{)}
\\&= \sum_{y \in \rangef{Y}} y \cd \sum_{x \in \rangef{X}} \P(Y=y \mid X=x,Z=z) \cd \P(X=x \mid Z=z) && \text{(exchange order of summation)}
\\&= \sum_{y \in \rangef{Y}} y \cd \P(Y=y \mid Z=z) && \text{(law of total probability given } Z=z \text{)}
\\&= \E{Y \mid Z=z} && \text{(definition of conditional expectation given } Z=z \text{)}
\ea
$$

**Continuous case.**
Conditioning on $Z=z$,
and integrating over $X$
under the conditional density $\p(X=x \mid Z=z)$:

$$
\ba
\E{\E{Y \mid X,Z} \mid Z=z}
&= \int_{x \in \rangef{X}} \E{Y \mid X=x,Z=z} \cd \p(X=x \mid Z=z)\, dx
\\&= \E{Y \mid Z=z}
\ea
$$

Therefore,
as random variables of $Z$,
$\E{Y \mid Z} = \E{\E{Y \mid X,Z} \mid Z}$.

:::

{{< slidebreak >}}

:::{#exm-lie}

#### Marginal expectation from conditional expectations

Suppose $X$ is a binary random variable indicating treatment assignment ($X=1$ treated, $X=0$ control),
with $\P(X=1) = 0.5$,
and suppose the outcome $Y$ has conditional expectations:

$$\E{Y \mid X=1} = 10, \quad \E{Y \mid X=0} = 6$$

By the law of iterated expectations (@thm-lie):

$$
\ba
\E{Y}
&= \E{\E{Y \mid X}}
\\&= \E{Y \mid X=1} \cd \P(X=1) + \E{Y \mid X=0} \cd \P(X=0)
\\&= 10 \cd 0.5 + 6 \cd 0.5
\\&= 5 + 3
\\&= 8
\ea
$$

:::
