::: remark

The sum of many independent random variables,
none of which dominates the others,
has a distribution that is approximately bell-shaped,
whatever the distributions of the individual variables.
@thm-clt makes this precise.

:::

{{< slidebreak >}}

:::{#thm-clt}
#### Central Limit Theorem

Let $X_1, X_2, \ldots$ be [IID](independence.qmd#def-iid) random variables
with mean $\mu$ and finite variance $\ss > 0$,
and let $S_n \eqdef \sumin X_i$.
Then for every real number $z$:

$$
\lim_{n \to \infty} \Pr\paren{\frac{S_n - n\mu}{\sigma\sqrt{n}} \le z} = \Phi(z)
$$

where $\Phi(z) \eqdef \int_{-\infty}^{z} \frac{1}{\sqrt{2\pi}} \text{e}^{-u^2/2}\,du$
is the CDF of the [standard normal distribution](random-variables.qmd#def-std-normal) $\ndist{0, 1}$.

:::

::: {.remark .notes}

This version is the Lindeberg--L&eacute;vy CLT;
its proof is beyond these notes' scope
[@billingsley1995probability, Theorem 27.1].
Other versions relax the IID assumption,
which is why the informal statement asks only that no summand dominate.

:::

{{< slidebreak >}}

:::{#cor-sum-iid-moments}
#### Mean and variance of a sum of IID random variables

Let $X_1, \ldots, X_n$ be [IID](independence.qmd#def-iid) random variables,
each discrete or continuous,
with mean $\mu$ and finite variance $\ss$,
and let $S_n \eqdef \sumin X_i$.
Then:

$$
\ba
\E{S_n} &= n\mu \\
\Var{S_n} &= n\ss
\ea
$$

:::

::: proof

For the mean, apply
[linearity of expectation](expectation.qmd#thm-linearity-expectation)
once per added summand:

$$
\ba
\E{S_n}
&= \sumin \E{X_i}
&& \text{(linearity of expectation, applied } n - 1 \text{ times)} \\
&= n\mu
&& \text{(each } X_i \text{ has mean } \mu \text{)}
\ea
$$

For the variance, apply the
[variance of a linear combination](variance-covariance.qmd#thm-var-lincom)
with every $a_i = 1$.
For $i \ne j$, $X_i$ and $X_j$ are independent
(take $A_k = \reals$ for every other $k$ in the
[definition of independence](independence.qmd#def-indpt)),
so $\Cov{X_i, X_j} = 0$ for
[independent summands](variance-covariance.qmd#thm-indpt-uncorrelated);
and $\Cov{X_i, X_i} = \Var{X_i}$
([covariance of a variable with itself](variance-covariance.qmd#lem-cov-xx)):

$$
\ba
\Var{S_n}
&= \sumin \sumn{j} \Cov{X_i, X_j}
&& \text{(variance of a linear combination, all } a_i = 1 \text{)} \\
&= \sumin \Cov{X_i, X_i} + \sum_{\substack{i, j = 1 \\ i \ne j}}^n \Cov{X_i, X_j}
&& \text{(split off the terms with } i = j \text{)} \\
&= \sumin \Var{X_i} + 0
&& \text{(covariance with itself; independent summands)} \\
&= n\ss
&& \text{(each } X_i \text{ has variance } \ss \text{)}
\ea
$$

:::

::: {.remark .notes}

In practice, @thm-clt justifies approximating
$S_n$ by a normal distribution with mean $n\mu$ and variance $n\ss$,
which by @cor-sum-iid-moments
are exactly the mean and variance of $S_n$.

:::

{{< slidebreak >}}

:::{#exm-clt-5d6}
#### The sum of five dice

A single fair die roll has the discrete uniform distribution on $\set{1, \ldots, 6}$,
which is flat, not bell-shaped (@fig-clt-1d6).
Its mean is $\mu = 3.5$,
and its variance is $\ss = \sum_{x=1}^{6} (x - 3.5)^2 / 6 = 35/12$.

```{r}
#| label: fig-clt-1d6
#| fig-cap: "Distribution of the outcome of one die"
#| code-fold: true
dice_sum_pmf <- function(n_dice) {
  totals <- rowSums(expand.grid(rep(list(1:6), n_dice)))
  probs <- prop.table(table(totals))
  data.frame(total = as.numeric(names(probs)), p = as.vector(probs))
}

dice_plot <-
  ggplot2::ggplot(mapping = ggplot2::aes(x = total, y = p)) +
  ggplot2::xlab("sum of dice (x)") +
  ggplot2::ylab("Probability of outcome, Pr(X=x)") +
  ggplot2::expand_limits(y = 0)

dice_plot + ggplot2::geom_col(data = dice_sum_pmf(1))
```

The sum of five independent rolls, $S_5$, is already close to bell-shaped
(@fig-clt-5d6).

```{r}
#| label: fig-clt-5d6
#| fig-cap: "Distribution of the sum of five dice"
#| code-fold: true
pmf_five_dice <- dice_sum_pmf(5)
dice_plot + ggplot2::geom_col(data = pmf_five_dice)
```

```{r}
#| include: false
p_exact <- sum(pmf_five_dice$p[pmf_five_dice$total <= 15])
p_normal <- pnorm(15.5, mean = 5 * 3.5, sd = sqrt(5 * 35 / 12))
```

For example, the exact probability that five dice total at most 15 is
$\Pr(S_5 \le 15) = `r round(p_exact, 4)`$.
The normal approximation that @thm-clt suggests,
with mean $5 \cd 3.5 = 17.5$ and variance $5 \cd 35/12 \approx 14.58$,
evaluated at $15.5$ to account for $S_5$ taking only integer values,
gives $\Phi\paren{(15.5 - 17.5)/\sqrt{14.58}} \approx `r round(p_normal, 4)`$.

:::
