::: notes

Some distributions are typically used for outcome models (@tbl-outcome-distns);
other distributions are typically used for test statistics (@tbl-test-stat-distns).

:::

:::{#tbl-outcome-distns}
Distribution | Uses
------------ | ---
Bernoulli | Binary outcomes
Binomial | Sums of Bernoulli outcomes
Poisson | Unbounded count outcomes
Geometric | Counts of non-events before an event occurs
Negative binomial | Mixtures of Poisson distributions, counts of non-events until a given number of events occurs
Normal (Gaussian) | Continuous outcomes without a more specific distribution
Exponential | Time to event outcomes
Gamma | Time to event outcomes
Weibull | Time to event outcomes
Log-normal | Time to event outcomes

Distributions typically used for outcome models
:::

{{< slidebreak >}}

:::{#tbl-test-stat-distns}
Distribution | Uses
------------ | ----
$\chi^2$ | Regression comparisons (asymptotic), contingency table independence tests, goodness-of-fit tests
$F$ | Gaussian model comparisons (exact)
$Z$ (standard normal) | Proportions, means, regression coefficients (asymptotic)
$t$ (Student's) | Means, regression coefficients in Gaussian outcome models (exact)

Distributions typically used for test statistics
:::

{{< slidebreak >}}

## The Bernoulli distribution {#sec-bern-dist}

:::{#def-bernoulli}
#### Bernoulli distribution

A random variable $X$ has the **Bernoulli distribution** with parameter $\pi \in [0, 1]$,
written $X \sim \Ber(\pi)$, if:

$$
\ba
\Pr(X=x) &\eqdef \1{x\in \set{0,1}}\pi^x(1-\pi)^{1-x}\\
&= \begin{cases}
\pi, & x=1\\
1-\pi, & x=0
\end{cases}
\ea
$$

:::

{{< slidebreak >}}

:::{#exm-bernoulli}
#### A coin flip

The indicator of heads on one flip of a fair coin is $\Ber(1/2)$.
By [earlier results](expectation.qmd#thm-bernoulli-mean),
$X \sim \Ber(\pi)$ has mean $\pi$
and [variance](variance-covariance.qmd#exm-variance-bernoulli) $\pi(1 - \pi)$.

:::

{{< slidebreak >}}

## The Poisson distribution {#sec-poisson-dist}

:::{#def-poisson}
#### Poisson distribution

A random variable $Y$ has the **Poisson distribution** with mean parameter $\mu > 0$,
written $Y \sim \Pois(\mu)$, if:

$$\P(Y = y) \eqdef \frac{\mu^{y} e^{-\mu}}{y!}, \quad y \in \Nat = \set{0, 1, 2, \dots}$$ {#eq-pois-pmf}

:::

::: notes
(see @fig-pois-pmf)
:::

{{< slidebreak >}}

:::{#exr-range-poisson}
What is the range of possible values for a Poisson distribution?
:::

{{< slidebreak >}}

:::{#sol-range-poisson}
$$\rangef{Y} = \set{0, 1, 2, \dots} = \Nat$$
:::

{{< slidebreak >}}

:::{#thm-cdf-pois}
#### CDF of Poisson distribution

$$\P(Y \le y) = e^{-\mu} \sum_{j=0}^{\floor{y}}\frac{\mu^j}{j!}$$ {#eq-pois-cdf}

:::

::: proof
For $y < 0$, both sides are $0$ (the sum is empty).
For any $y \ge 0$,
the event $\{Y \le y\}$ is the disjoint union of events $\{Y = j\}$
for all non-negative integers $j \le \floor{y}$.
Applying the Poisson PMF (@eq-pois-pmf) and factoring out the common term $e^{-\mu}$:

$$
\ba
\P(Y \le y)
&= \sum_{j=0}^{\floor{y}} \P(Y = j) && (\text{disjoint union of events } Y = j) \\
&= \sum_{j=0}^{\floor{y}} \frac{\mu^j e^{-\mu}}{j!} && (\text{definition of Poisson PMF}) \\
&= e^{-\mu} \sum_{j=0}^{\floor{y}} \frac{\mu^j}{j!} && (\text{factoring out } e^{-\mu} \text{ constant wrt } j)
\ea
$$
:::

{{< slidebreak >}}

:::{#exm-cdf-pois}
#### Computing Poisson cumulative probabilities

For a Poisson random variable $X \sim \Pois(\mu = 2)$,
the probability of observing at most 2 events is computed as:

$$
\ba
\P(X \le 2)
&= e^{-2} \sum_{j=0}^{2} \frac{2^j}{j!} && (\text{apply CDF formula with } \mu = 2, y = 2) \\
&= e^{-2} \paren{\frac{2^0}{0!} + \frac{2^1}{1!} + \frac{2^2}{2!}} && (\text{expand terms for } j = 0, 1, 2) \\
&= e^{-2} \paren{1 + 2 + 2} && (\text{simplify factorials and powers}) \\
&= 5 e^{-2} \approx 0.677 && (\text{evaluate numerical value})
\ea
$$
:::

::: notes
(see @fig-pois-cdfs)
:::

{{< slidebreak >}}

```{r}
#| label: fig-pois-pmf
#| fig-cap: "Poisson PMFs, by mean parameter $\\mu$"
#| code-fold: true

pois_dists <-
  dplyr::tibble(mu = c(0.5, 1, 2, 5, 10, 20)) |>
  dplyr::reframe(.by = mu, x = 0:30) |>
  dplyr::mutate(
    `P(X = x)` = dpois(x, lambda = mu),
    `P(X <= x)` = ppois(x, lambda = mu),
    mu = factor(mu)
  )

plot0 <-
  pois_dists |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = x,
      y = `P(X = x)`,
      fill = mu,
      col = mu
    )
  ) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    fill = latex2exp::TeX("$\\mu$"),
    col = latex2exp::TeX("$\\mu$"),
    y = latex2exp::TeX("$\\Pr_{\\mu}(X = x)$")
  )

plot1 <-
  plot0 +
  ggplot2::geom_segment(yend = 0) +
  ggplot2::facet_wrap(~mu)

print(plot1)
```

{{< slidebreak >}}

```{r}
#| label: fig-pois-cdfs
#| fig-cap: "Poisson CDFs"
#| code-fold: true

plot2 <-
  plot0 +
  ggplot2::geom_step(alpha = 0.75) +
  ggplot2::aes(y = `P(X <= x)`) +
  ggplot2::labs(y = latex2exp::TeX("$\\Pr_{\\mu}(X \\leq x)$"))

print(plot2)
```

{{< slidebreak >}}

:::{#exr-pois-dist-funs}
#### Poisson distribution functions

```{r}
#| label: def-mu-exr-pois
#| include: false
mu <- 3.75
```

Let $X \sim \Pois(\mu = `r mu`)$.

Compute:

- $\P(X = 4 | \mu = `r mu`)$
- $\P(X \le 7 | \mu = `r mu`)$
- $\P(X > 5 | \mu = `r mu`)$

:::

{{< slidebreak >}}

::: solution

- $\P(X=4) = `r dpois(x = 4, lambda = mu)`$
- $\P(X\le 7) = `r ppois(q = 7, lambda = mu)`$
- $\P(X > 5) = `r ppois(q = 5, lambda = mu, lower.tail = FALSE)`$

:::

{{< slidebreak >}}

:::{#thm-poisson-properties}
#### Properties of the Poisson distribution

If $X \sim \Pois(\mu)$, then:

* $\Expp[X] = \mu$
* $\Varr(X) = \mu$
* $\P(X=x) = \frac{\mu}{x} \P(X = x-1)$ for $x \in \set{1, 2, \dots}$
* For $x \in \set{1, 2, \dots}$ with $x < \mu$, $\P(X=x) > \P(X = x-1)$
* For $x = \mu$ (possible only when $\mu$ is an integer), $\P(X=x) = \P(X = x-1)$
* For $x \in \set{1, 2, \dots}$ with $x > \mu$, $\P(X=x) < \P(X = x-1)$
* If $\mu$ is not an integer, $\arg \max_{x} \P(X=x) = \floor{\mu}$;
  if $\mu$ is an integer, the maximum is attained at both $x = \mu - 1$ and $x = \mu$

:::

::: proof

**Mean.**

$$
\ba
\Expp[X]
&= \sum_{x=0}^\infty x \cdot \P(X=x) && (\text{definition of expected value}) \\
&= 0 \cdot \P(X=0) + \sum_{x=1}^\infty x \cdot \P(X=x) && (\text{separate } x=0 \text{ term}) \\
&= \sum_{x=1}^\infty x \cdot \frac{\mu^x e^{-\mu}}{x!} && (\text{substitute Poisson PMF}) \\
&= \sum_{x=1}^\infty x \cdot \frac{\mu^x e^{-\mu}}{x \cdot (x-1)!} && (\text{definition of factorial } x!) \\
&= \sum_{x=1}^\infty \frac{\mu^x e^{-\mu}}{(x-1)!} && (\text{cancel factor of } x) \\
&= \mu \cdot \sum_{x=1}^\infty \frac{\mu^{x-1} e^{-\mu}}{(x-1)!} && (\text{factor out one power of } \mu) \\
&= \mu \cdot \sum_{y=0}^\infty \frac{\mu^y e^{-\mu}}{y!} && (\text{change index variable } y \eqdef x-1) \\
&= \mu \cdot 1 && (\text{PMF sums to 1 over state space}) \\
&= \mu && (\text{simplify})
\ea
$$

**Variance.**
The same steps, canceling two factors instead of one, give $\E{X(X-1)}$:

$$
\ba
\E{X(X-1)}
&= \sum_{x=2}^\infty x(x-1) \cdot \frac{\mu^x e^{-\mu}}{x!} && (\text{LOTUS; the } x = 0, 1 \text{ terms are } 0) \\
&= \sum_{x=2}^\infty \frac{\mu^x e^{-\mu}}{(x-2)!} && (\text{cancel } x(x-1) \text{ against } x!) \\
&= \mu^2 \cdot \sum_{y=0}^\infty \frac{\mu^y e^{-\mu}}{y!} && (\text{factor out } \mu^2 \text{; } y \eqdef x - 2) \\
&= \mu^2 && (\text{PMF sums to 1})
\ea
$$

Then, by the [simplified expression for variance](variance-covariance.qmd#thm-variance)
and [linearity of expectation](expectation.qmd#thm-linearity-expectation):

$$
\ba
\Varr(X)
&= \E{X^2} - \sqf{\E{X}} && (\text{simplified expression for variance}) \\
&= \E{X(X-1)} + \E{X} - \sqf{\E{X}} && (X^2 = X(X-1) + X \text{; linearity}) \\
&= \mu^2 + \mu - \mu^2 && (\text{substitute}) \\
&= \mu && (\text{simplify})
\ea
$$

**Ratio of consecutive probabilities.**
For $x \in \set{1, 2, \dots}$:

$$
\ba
\frac{\P(X = x)}{\P(X = x - 1)}
&= \frac{\mu^x e^{-\mu} / x!}{\mu^{x-1} e^{-\mu} / (x-1)!} && (\text{substitute Poisson PMF}) \\
&= \frac{\mu}{x} && (\text{cancel } \mu^{x-1} e^{-\mu} \text{ and } (x-1)!)
\ea
$$

This ratio is greater than, equal to, or less than 1
as $x$ is less than, equal to, or greater than $\mu$,
which gives the three comparisons.
So $\P(X = x)$ increases while $x < \mu$ and decreases once $x > \mu$.
If $\mu$ is not an integer, the last increase is at $x = \floor{\mu}$,
which is therefore the unique mode.
If $\mu$ is an integer, $\P(X = \mu) = \P(X = \mu - 1)$,
and both are modes.

See also <https://statproofbook.github.io/P/poiss-mean>
and <https://statproofbook.github.io/P/poiss-var>.

:::

{{< slidebreak >}}

:::{#exm-poisson-properties}
#### Mode of a Poisson distribution

For $X \sim \Pois(2.5)$,
$\P(X = 2) = \frac{2.5}{2}\P(X = 1) > \P(X = 1)$
and $\P(X = 3) = \frac{2.5}{3}\P(X = 2) < \P(X = 2)$,
so the mode is $\floor{2.5} = 2$.
For $X \sim \Pois(2)$, $\P(X = 2) = \frac{2}{2}\P(X = 1)$,
so $1$ and $2$ are both modes,
each with probability $2 e^{-2} \approx 0.271$.

:::

{{< slidebreak >}}

:::{#def-exposure}
#### Exposure magnitude

For many count outcomes,
there is some sense of
an **exposure magnitude**,
such as
**population size** or
**duration of observation**,
which multiplicatively rescales
the expected (mean) count.

:::

{{< slidebreak >}}

:::{#exr-exposure-magnitude}
What are some examples of exposure magnitudes?
:::

{{< slidebreak >}}

::: {.solution .smaller}

outcome                      | exposure units
-----------------------------| -------------
disease incidence            | number of individuals exposed; time at risk
car accidents                | miles driven
worksite accidents           | person-hours worked
population size              | size of habitat

: Examples of exposure units {#tbl-exposure-units}

:::

:::: notes
Exposure units are similar to
the number of trials in a binomial distribution,
but **in non-binomial count outcomes, there can be more than one event per unit of exposure**.

We can use $t$ to represent continuous-valued exposures/observation durations,
and $n$ to represent discrete-valued exposures.

::::

{{< slidebreak >}}

:::{#def-event-rate}
#### Event rate

For a count outcome $Y$ with exposure magnitude $t$,
the **event rate** (denoted $\lambda$)
is defined as the mean of $Y$ divided by the exposure magnitude:

$$\mu \eqdef \Expp[Y \mid T=t]$$

$$\lambda \eqdef \frac{\mu}{t}$$ {#eq-def-event-rate}

:::

::: notes
Event rate is somewhat analogous to odds for a binary outcome:
both are transformations of the mean.
The event rate removes the exposure magnitude from the mean,
so counts observed over different exposures can be compared on one scale.
In count regression models, the transformation $\lambda = \mu/t$ is not part of the model's link function,
and the exposure magnitude is handled differently from the other covariates
(see [rme's count-regression chapter](https://morrison-lab.github.io/rme/chapters/count-regression.html)).
:::

{{< slidebreak >}}

:::{#thm-mean-vs-event-rate}
#### Transformation function from event rate to mean

For a count variable with mean $\mu$, event rate $\lambda$, and exposure magnitude $t > 0$:

$$\mu = \lambda \cdot t$$ {#eq-lambda-to-mu}

:::

::: proof
By @def-event-rate:

$$
\ba
\lambda &\eqdef \frac{\mu}{t} && (\text{definition of event rate}) \\
\mu &= \lambda \cdot t && (\text{multiply both sides by } t > 0)
\ea
$$
:::

{{< slidebreak >}}

:::{#exm-mean-vs-event-rate}
#### Calculating expected counts from event rates

Suppose a city records a disease event rate of $\lambda = 0.05$ cases per person-year.
For a subpopulation with an exposure magnitude of $t = 100$ person-years,
the expected count of cases is, by @thm-mean-vs-event-rate:

$$
\ba
\mu &= \lambda \cdot t && (\text{transformation from event rate to mean}) \\
&= 0.05 \times 100 && (\text{substitute } \lambda = 0.05 \text{ and } t = 100) \\
&= 5 \text{ cases} && (\text{evaluate expected count})
\ea
$$
:::

::: notes

@eq-lambda-to-mu is analogous to the inverse-odds function for binary variables.

:::

{{< slidebreak >}}

:::{#thm-non-exposed}
#### No exposure means no expected events

If the mean count is proportional to the exposure magnitude,
$\E{Y \mid T=t} = \lambda \cdot t$ for all $t \ge 0$ with one finite rate $\lambda$,
then there are no expected events without exposure:

$$\Expp[Y \mid T=0] = 0$$

:::

::: proof
$$
\ba
\Expp[Y \mid T=0]
&= \lambda \cdot 0 && (\text{evaluate } \E{Y \mid T = t} = \lambda t \text{ at } t = 0) \\
&= 0 && (\text{multiplication by zero; } \lambda \text{ is finite})
\ea
$$
:::

::: notes

The hypothesis carries the content here.
@def-event-rate alone cannot give $\E{Y \mid T = 0} = 0$,
since $\lambda = \mu / t$ is undefined at $t = 0$;
the result holds for a model that assumes a rate $\lambda$
shared across exposure magnitudes, including $t = 0$.

:::

{{< slidebreak >}}

:::{#exm-non-exposed}
#### Zero exposure time

If a subject is observed for $t = 0$ person-years,
no follow-up time has elapsed,
so under a constant-rate model
the expected number of incident events is $\Expp[Y \mid T=0] = 0$.
:::

{{< slidebreak >}}

:::{.callout-important}

The exposure magnitude, $T$, is *similar* to a covariate in linear or logistic regression.
However, there is an important difference: in count regression, **there is no intercept corresponding to $\Expp[Y|T=0]$**.
In other words, this model assumes that if there is no exposure, there can't be any events.

:::

{{< slidebreak >}}

:::{#thm-exposure-log-scale}
#### Exposure is additive on the log scale

If $\mu = \lambda\cdot t$ with $\lambda > 0$ and $t > 0$, then:

$$\log{\mu} = \log{\lambda} + \log{t}$$

:::

::: proof
$$
\ba
\log{\mu}
&= \log(\lambda \cdot t) && (\text{substitute } \mu = \lambda \cdot t) \\
&= \log{\lambda} + \log{t} && (\text{logarithm product rule})
\ea
$$
:::

{{< slidebreak >}}

:::{#exm-exposure-log-scale}
#### Log-linear representation of expected counts

If a clinic sees an event rate of $\lambda = 0.02$ events/day and $t = 30$ days of observation:

$$
\ba
\log{\mu}
&= \log(0.02) + \log(30) && (\text{apply log-scale formula}) \\
&= -3.912 + 3.401 && (\text{evaluate natural logarithms}) \\
&= -0.511 && (\text{sum terms})
\ea
$$

Exponentiating yields $\mu = \expf{-0.511} \approx 0.60$ expected events.
:::

{{< slidebreak >}}

:::{#def-offset}
#### Offset

For a count outcome with exposure magnitude $t$,
the known term $\log{t}$ in the log-scale decomposition of the mean
([@thm-exposure-log-scale]),

$$\log{\mu} = \log{\lambda} + \log{t},$$

is called the **offset**:
it shifts $\log{\mu}$ by a known amount,
with no unknown coefficient to estimate.

:::

::: notes
The offset needs no covariates:
with a single exposure $t$ and an unknown rate $\lambda$,
$\log{t}$ is already an offset.
Regression models for counts keep the same term
and add covariate terms beside it
(see [rme's count-regression chapter](https://morrison-lab.github.io/rme/chapters/count-regression.html)).

:::

{{< slidebreak >}}

:::{#exm-offset}
#### The offset for a clinic's event count

In @exm-exposure-log-scale,
the clinic was observed for $t = 30$ days,
so the offset is $\log{30} \approx 3.401$.
Only $\log{\lambda}$ is unknown before the data are seen;
the offset is fixed by the length of observation.
A clinic observed for $t = 60$ days has offset $\log{60} \approx 4.094$,
which raises $\log{\mu}$ by $\log{2} \approx 0.693$ at the same event rate.
:::

{{< slidebreak >}}

:::{#thm-sum-pois}
#### Sum of independent Poisson random variables

If $X$ and $Y$ are [independent](independence.qmd#def-indpt) Poisson random variables with means
$\mu_X$ and $\mu_Y$, their sum, $Z=X+Y$, is also a Poisson random variable, with mean 
$\mu_Z = \mu_X + \mu_Y$.

:::

::: proof
The event $\{Z = z\}$ is the disjoint union of the events $\{X = k, Y = z - k\}$
for $k = 0, 1, \ldots, z$
(since $X$ and $Y$ are non-negative integers), so:

$$
\ba
\P(Z = z)
&= \sum_{k=0}^z \P(X = k,\, Y = z - k) && (\text{countable additivity}) \\
&= \sum_{k=0}^z \P(X = k) \P(Y = z - k) && (\text{independence}) \\
&= \sum_{k=0}^z \frac{\mu_X^k e^{-\mu_X}}{k!} \frac{\mu_Y^{z-k} e^{-\mu_Y}}{(z-k)!} && (\text{substitute Poisson PMFs}) \\
&= e^{-(\mu_X + \mu_Y)} \sum_{k=0}^z \frac{\mu_X^k \mu_Y^{z-k}}{k!(z-k)!} && (\text{factor out } e^{-\mu_X} e^{-\mu_Y}) \\
&= \frac{e^{-(\mu_X + \mu_Y)}}{z!} \sum_{k=0}^z \frac{z!}{k!(z-k)!} \mu_X^k \mu_Y^{z-k} && (\text{multiply and divide by } z!) \\
&= \frac{e^{-(\mu_X + \mu_Y)}}{z!} (\mu_X + \mu_Y)^z && (\text{binomial theorem})
\ea
$$

This probability expression matches the PMF of a $\Pois(\mu_X + \mu_Y)$ random variable
(see also <https://web.stanford.edu/class/archive/cs/cs109/cs109.1206/lectureNotes/LN12_independent_rvs.pdf>, Example 3).
:::

{{< slidebreak >}}

:::{#exm-sum-pois}
#### Aggregating independent region counts

Suppose Region A records $X \sim \Pois(\mu_X = 12)$ cases
and Region B records $Y \sim \Pois(\mu_Y = 18)$ cases independently.
By @thm-sum-pois, the combined total count $Z = X + Y$ follows a Poisson distribution:

$$
\ba
Z &\sim \Pois(\mu_X + \mu_Y) && (\text{sum of independent Poissons}) \\
&= \Pois(12 + 18) && (\text{substitute region means}) \\
&= \Pois(30) && (\text{evaluate sum})
\ea
$$
:::

{{< slidebreak >}}

## The Negative-Binomial distribution {#sec-nb-dist}

:::{#def-nb}
#### Negative binomial distribution

A random variable $Y$ has the **negative binomial distribution**
with mean $\mu > 0$ and overdispersion parameter $\rho > 0$,
written $Y \sim \NegBin(\mu, \rho)$, if, for $y \in \set{0, 1, 2, \dots}$:

$$
\P(Y=y)
\eqdef \frac{\mu^y}{y!}
\cdot \frac{\Gamma(\rho + y)}{\Gamma(\rho) \cdot (\rho + \mu)^y}
\cdot \left(1+\frac{\mu}{\rho}\right)^{-\rho}
$$

where $\Gamma$ is the gamma function,
which satisfies $\Gamma(x) = (x-1)!$ for positive integers $x$.

:::

::: notes
As $\rho \rightarrow \infty$,
the second factor converges to 1
and the third factor converges to $\expf{-\mu}$,
which brings us back to the Poisson distribution.
:::

{{< slidebreak >}}

:::{#thm-nb}
#### Mean and variance of the negative binomial distribution

If $Y \sim \NegBin(\mu, \rho)$, then:

- $\Expp[Y] = \mu$
- $\Var{Y} = \mu + \frac{\mu^2}{\rho} > \mu$

:::

::: proof

The negative binomial distribution is a gamma mixture of Poisson distributions:
if $\Lambda$ has the gamma density
$g(\lambda) = \frac{(\rho/\mu)^\rho}{\Gamma(\rho)} \lambda^{\rho - 1} e^{-\rho\lambda/\mu}$
for $\lambda > 0$,
which has mean $\mu$ and variance $\mu^2/\rho$ [@CaseBerg01],
and $Y \mid \Lambda = \lambda \sim \Pois(\lambda)$,
then $Y \sim \NegBin(\mu, \rho)$.
To check this, integrate the joint density over $\lambda$:

$$
\ba
\P(Y = y)
&= \int_0^\infty \frac{\lambda^y e^{-\lambda}}{y!} \cd \frac{(\rho/\mu)^\rho}{\Gamma(\rho)} \lambda^{\rho - 1} e^{-\rho\lambda/\mu}\,d\lambda
&& (\text{marginalize over } \Lambda) \\
&= \frac{(\rho/\mu)^\rho}{y!\,\Gamma(\rho)} \int_0^\infty \lambda^{y + \rho - 1} e^{-\lambda(1 + \rho/\mu)}\,d\lambda
&& (\text{collect powers of } \lambda \text{ and exponents}) \\
&= \frac{(\rho/\mu)^\rho}{y!\,\Gamma(\rho)} \cd \frac{\Gamma(y + \rho)}{(1 + \rho/\mu)^{y + \rho}}
&& (\textstyle\int_0^\infty \lambda^{a-1} e^{-b\lambda}\,d\lambda = \Gamma(a)/b^a) \\
&= \frac{\Gamma(y + \rho)}{y!\,\Gamma(\rho)} \paren{\frac{\rho}{\mu}}^\rho \paren{\frac{\mu}{\mu + \rho}}^{y + \rho}
&& (1 + \rho/\mu = (\mu + \rho)/\mu) \\
&= \frac{\Gamma(y + \rho)}{y!\,\Gamma(\rho)} \paren{\frac{\rho}{\mu + \rho}}^{\rho} \paren{\frac{\mu}{\mu + \rho}}^{y}
&& (\text{combine the } \mu^\rho \text{ factors}) \\
&= \frac{\mu^y}{y!} \cd \frac{\Gamma(\rho + y)}{\Gamma(\rho)\,(\rho + \mu)^y} \cd \paren{1 + \frac{\mu}{\rho}}^{-\rho}
&& (\text{rearrange into the form of the definition})
\ea
$$

Then, by the [law of iterated expectations](expectation.qmd#thm-lie)
and the [law of total variance](variance-covariance.qmd#thm-total-variance),
using $\E{Y \mid \Lambda} = \Varr(Y \mid \Lambda) = \Lambda$
(@thm-poisson-properties):

$$
\ba
\Expp[Y] &= \E{\E{Y \mid \Lambda}} = \E{\Lambda} = \mu \\
\Var{Y} &= \E{\Var{Y \mid \Lambda}} + \Var{\E{Y \mid \Lambda}}
= \E{\Lambda} + \Var{\Lambda}
= \mu + \frac{\mu^2}{\rho}
\ea
$$

and $\mu^2/\rho > 0$ gives $\Var{Y} > \mu$.

:::

{{< slidebreak >}}

:::{#exm-nb}
#### Overdispersion relative to the Poisson

With $\mu = 4$ and $\rho = 2$,
$\Var{Y} = 4 + 16/2 = 12$,
three times the variance of a $\Pois(4)$ count with the same mean.

:::

{{< slidebreak >}}

## Weibull distribution {#sec-weibull}

:::{#def-weibull}
#### Weibull distribution

A non-negative random variable $T$ has the **Weibull distribution**
with shape $\alpha > 0$ and rate $\lambda > 0$ if its
[survival function](random-variables.qmd#def-surv-fn) is:

$$\surv(t) \eqdef \text{e}^{-\lambda t^\alpha}, \quad t \ge 0$$

:::

{{< slidebreak >}}

:::{#thm-weibull}
#### Weibull density, hazard, and mean

If $T$ has the Weibull distribution with shape $\alpha$ and rate $\lambda$, then for $t > 0$:

$$
\begin{aligned}
f(t) &= \alpha\lambda t^{\alpha-1}\text{e}^{-\lambda t^\alpha}\\
\h(t) &= \alpha\lambda t^{\alpha-1}\\
\E{T} &= \Gamma(1+1/\alpha)\cdot \lambda^{-1/\alpha}
\end{aligned}
$$

:::

::: proof

The CDF of $T$ is $F(t) = 1 - \surv(t)$
([survival function and CDF](random-variables.qmd#thm-survival-expressions-1)),
which is $0$ for $t < 0$ and $1 - \text{e}^{-\lambda t^\alpha}$ for $t \ge 0$.
This $F$ is continuous everywhere,
with a continuous derivative everywhere except possibly $t = 0$,
so $f = F' = -\surv'$ is a density of $T$
([a piecewise-smooth CDF has its derivative as a density](random-variables.qmd#thm-cdf-derivative-density)).
This density is continuous at every $t > 0$,
so there the hazard is $f(t)/\surv(t)$
([hazard equals density over survival](random-variables.qmd#thm-hazard-dens-surv)):

$$
\ba
f(t)
&= -\frac{d}{dt}\text{e}^{-\lambda t^\alpha}
&& (f = -\surv') \\
&= \alpha\lambda t^{\alpha-1}\text{e}^{-\lambda t^\alpha}
&& (\text{chain rule}) \\
\h(t)
&= \frac{\alpha\lambda t^{\alpha-1}\text{e}^{-\lambda t^\alpha}}{\text{e}^{-\lambda t^\alpha}}
&& (\text{hazard is density over survival}) \\
&= \alpha\lambda t^{\alpha-1}
&& (\text{cancel})
\ea
$$

For the mean, use the [survival-function formula for the mean](expectation.qmd#thm-surv-mean)
and substitute $u = \lambda t^\alpha$,
so that $t = (u/\lambda)^{1/\alpha}$ and $dt = \frac{1}{\alpha}\lambda^{-1/\alpha} u^{1/\alpha - 1}\,du$:

$$
\ba
\E{T}
&= \int_0^\infty \text{e}^{-\lambda t^\alpha}\,dt
&& (\text{expectation via the survival function}) \\
&= \frac{1}{\alpha}\lambda^{-1/\alpha} \int_0^\infty u^{1/\alpha - 1} \text{e}^{-u}\,du
&& (\text{substitute } u = \lambda t^\alpha) \\
&= \frac{1}{\alpha}\lambda^{-1/\alpha}\,\Gamma(1/\alpha)
&& (\text{definition of the gamma function}) \\
&= \Gamma(1 + 1/\alpha)\,\lambda^{-1/\alpha}
&& (\Gamma(1 + a) = a\,\Gamma(a))
\ea
$$

:::

::: notes

The hazard is written $\h(t)$ here, rather than $\haz(t)$,
because the Weibull rate parameter is also called $\lambda$.

When $\alpha=1$, the Weibull distribution reduces to the exponential distribution.
When $\alpha>1$ the hazard is
increasing and when $\alpha < 1$ the hazard is decreasing.
This property provides
more flexibility than the exponential.

:::

{{< slidebreak >}}

:::{#exm-weibull}
#### Exponential as a special case

With $\alpha = 1$, @thm-weibull gives $\h(t) = \lambda$ and
$\E{T} = \Gamma(2)\lambda^{-1} = 1/\lambda$,
matching the exponential distribution's constant hazard
and mean.
With $\alpha = 2$ and $\lambda = 1$,
$\h(t) = 2t$ increases with $t$,
and $\E{T} = \Gamma(3/2) = \sqrt{\pi}/2 \approx 0.886$.

:::
